---
title: "DASC 460: Analyzing the Spatial Spread of Violence in Saint Paul to Determine the Effectiveness of Establishing a Non-Fatal Shooting Unit"
output: html_document
---

------------------------------------------------------------------------

# Setup

## Load Libraries

```{r warning=FALSE, message=FALSE}
library(rmarkdown)
library(knitr)
library(igraph)
library(corrplot)
library(moments)
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(readr)
library(shiny)
library(geosphere)
```

## Load Data

```{r}
file <- "Shootings_Full Data(new).xlsx"
excel_sheets(file)

full_data <- read_excel(file, sheet = "Full Data")
loc_time  <- read_excel(file, sheet = "Loc_Time")
tables    <- read_excel(file, sheet = "Tables")
```

## Examine Data

```{r}
names(full_data)
names(loc_time)
names(tables)

head(full_data)
head(loc_time)
head(tables)
```

## Clean Data

```{r}
# Keep only rows with valid coordinates
map_data <- full_data %>%
  filter(!is.na(Lat), !is.na(Long), !is.na(DATE), Lat != 0, Long != 0) %>%
  mutate(
    DATE      = as.Date(DATE),
    Year      = format(DATE, "%Y"),
    month_day = format(DATE, "%m-%d")
  )
```

## Set Time Variables

```{r}
all_dates <- sort(unique(map_data$DATE))
years     <- sort(unique(map_data$Year))
years4    <- years[1:min(4, length(years))]
```

------------------------------------------------------------------------

# Section 1: Interactive Maps

## Visualization 1 — Point Map

Individual shooting incidents plotted as circle markers.

```{r}
leaflet(map_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~Long,
    lat         = ~Lat,
    radius      = 3,
    color       = "purple",
    opacity     = 0.8,
    fillOpacity = 0.4,
    popup = ~paste(
      "<b>CN:</b>", CN, "<br>",
      "<b>Date:</b>", as.Date(DATE), "<br>",
      "<b>Description:</b>", DESCRIPTION, "<br>",
      "<b>Disposition:</b>", DISPOSITION
    )
  )
```

## Visualization 2 — Cluster Aggregation Map

Markers grouped into clusters to show density by area.

```{r}
leaflet(map_data) %>%
  addTiles() %>%
  addMarkers(
    lng           = ~Long,
    lat           = ~Lat,
    clusterOptions = markerClusterOptions(showCoverageOnHover = TRUE)
  )
```

## Visualization 3 — Grid Aggregation Map

Circle size reflects the number of incidents within each grid cell.

```{r}
grid_counts <- full_data %>%
  filter(!is.na(GRID), !is.na(Lat), !is.na(Long)) %>%
  group_by(GRID) %>%
  summarize(
    count = n(),
    Lat   = mean(Lat),
    Long  = mean(Long)
  )

leaflet(grid_counts) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~Long,
    lat         = ~Lat,
    radius      = ~sqrt(count) * 2,
    color       = "red",
    fillOpacity = 0.5,
    popup = ~paste(
      "<b>GRID:</b>", GRID, "<br>",
      "<b>Count:</b>", count
    )
  )
```

## Visualization 4 — Animated Small Multiples (Shiny App)

Interactive Shiny app showing shooting progression across four years simultaneously. Use the slider or Play/Pause buttons to animate through the calendar year.

```{r}
# Use leap-year style full-year sequence so all month/day values exist
slider_days   <- seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day")
slider_labels <- format(slider_days, "%m-%d")

make_base_map <- function(df) {
  leaflet() %>%
    addTiles() %>%
    setView(
      lng  = mean(df$Long, na.rm = TRUE),
      lat  = mean(df$Lat,  na.rm = TRUE),
      zoom = 11
    )
}

ui <- fluidPage(
  titlePanel("Shootings by Year: Synchronized Time Progression"),

  fluidRow(
    column(3,
      tags$b("Progress through year:"),
      textOutput("date_label")
    ),
    column(3,
      actionButton("play",  "Play"),
      actionButton("pause", "Pause")
    ),
    column(6,
      sliderInput("speed", "Animation speed (smaller = faster)",
                  min = 100, max = 2000, value = 700, step = 100)
    )
  ),

  sliderInput(
    inputId = "day_index",
    label   = "Move through the calendar year",
    min     = 1,
    max     = length(slider_labels),
    value   = 1,
    step    = 1,
    animate = FALSE
  ),

  tabsetPanel(
    tabPanel("Points",
      fluidRow(
        column(6, h4(if (length(years4) >= 1) years4[1] else ""), leafletOutput("points_1", height = 325)),
        column(6, h4(if (length(years4) >= 2) years4[2] else ""), leafletOutput("points_2", height = 325))
      ),
      fluidRow(
        column(6, h4(if (length(years4) >= 3) years4[3] else ""), leafletOutput("points_3", height = 325)),
        column(6, h4(if (length(years4) >= 4) years4[4] else ""), leafletOutput("points_4", height = 325))
      )
    ),
    tabPanel("Clusters",
      fluidRow(
        column(6, h4(if (length(years4) >= 1) years4[1] else ""), leafletOutput("clusters_1", height = 325)),
        column(6, h4(if (length(years4) >= 2) years4[2] else ""), leafletOutput("clusters_2", height = 325))
      ),
      fluidRow(
        column(6, h4(if (length(years4) >= 3) years4[3] else ""), leafletOutput("clusters_3", height = 325)),
        column(6, h4(if (length(years4) >= 4) years4[4] else ""), leafletOutput("clusters_4", height = 325))
      )
    )
  )
)

server <- function(input, output, session) {

  playing <- reactiveVal(FALSE)

  observeEvent(input$play,  { playing(TRUE)  })
  observeEvent(input$pause, { playing(FALSE) })

  observe({
    if (playing()) {
      invalidateLater(input$speed, session)
      current <- input$day_index
      if (current < length(slider_labels)) {
        updateSliderInput(session, "day_index", value = current + 1)
      } else {
        playing(FALSE)
      }
    }
  })

  selected_month_day <- reactive({ slider_labels[input$day_index] })

  output$date_label <- renderText({ selected_month_day() })

  # Initialize blank base maps
  output$points_1   <- renderLeaflet({ make_base_map(map_data) })
  output$points_2   <- renderLeaflet({ make_base_map(map_data) })
  output$points_3   <- renderLeaflet({ make_base_map(map_data) })
  output$points_4   <- renderLeaflet({ make_base_map(map_data) })
  output$clusters_1 <- renderLeaflet({ make_base_map(map_data) })
  output$clusters_2 <- renderLeaflet({ make_base_map(map_data) })
  output$clusters_3 <- renderLeaflet({ make_base_map(map_data) })
  output$clusters_4 <- renderLeaflet({ make_base_map(map_data) })

  # Helper: filter a year up to the selected month-day
  filtered_year_data <- function(yr, month_day_cutoff) {
    map_data %>% filter(Year == yr, month_day <= month_day_cutoff)
  }

  # Points small multiples
  observe({
    for (i in seq_along(years4)) {
      yr      <- years4[i]
      df_year <- filtered_year_data(yr, selected_month_day())

      leafletProxy(paste0("points_", i), data = df_year) %>%
        clearMarkers() %>%
        clearMarkerClusters() %>%
        addCircleMarkers(
          lng         = ~Long,
          lat         = ~Lat,
          radius      = 3,
          color       = "red",
          opacity     = 0.8,
          fillOpacity = 0.4,
          popup = ~paste(
            "<b>Year:</b>", Year, "<br>",
            "<b>CN:</b>", CN, "<br>",
            "<b>Date:</b>", DATE, "<br>",
            "<b>Description:</b>", DESCRIPTION, "<br>",
            "<b>Disposition:</b>", DISPOSITION
          )
        )
    }
  })

  # Clusters small multiples
  observe({
    for (i in seq_along(years4)) {
      yr      <- years4[i]
      df_year <- filtered_year_data(yr, selected_month_day())

      leafletProxy(paste0("clusters_", i), data = df_year) %>%
        clearMarkers() %>%
        clearMarkerClusters() %>%
        addMarkers(
          lng  = ~Long,
          lat  = ~Lat,
          popup = ~paste(
            "<b>Year:</b>", Year, "<br>",
            "<b>CN:</b>", CN, "<br>",
            "<b>Date:</b>", DATE, "<br>",
            "<b>Description:</b>", DESCRIPTION, "<br>",
            "<b>Disposition:</b>", DISPOSITION
          ),
          clusterOptions = markerClusterOptions(showCoverageOnHover = TRUE)
        )
    }
  })

  # Clear unused panels when fewer than 4 years present
  observe({
    if (length(years4) < 4) {
      for (i in (length(years4) + 1):4) {
        leafletProxy(paste0("points_",   i)) %>% clearMarkers() %>% clearMarkerClusters()
        leafletProxy(paste0("clusters_", i)) %>% clearMarkers() %>% clearMarkerClusters()
      }
    }
  })
}

shinyApp(ui, server)
```

------------------------------------------------------------------------

# Section 2: Temporal Patterns

## Prepare Temporal Variables

```{r}
map_data <- map_data %>%
  mutate(
    month  = floor_date(DATE, "month"),
    period = ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
    hour   = hour(Time),
    day    = wday(DATE, label = TRUE, week_start = 1)
  ) %>%
  mutate(
    period = factor(period, levels = c("Before NFSU", "After NFSU")),
    day    = factor(day,    levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
  )
```

## Visualization 5.1 — Monthly Counts (Line)

```{r}
monthly_counts <- map_data %>%
  group_by(month) %>%
  summarize(count = n(), .groups = "drop")

ggplot(monthly_counts, aes(x = month, y = count)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  labs(title = "Monthly Shootings Over Time", x = "Month", y = "Count")
```

```{r}
monthly_period_counts <- map_data %>%
  group_by(period, month) %>%
  summarize(count = n(), .groups = "drop")

ggplot(monthly_period_counts, aes(x = month, y = count)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  facet_wrap(~period, scales = "fixed") +
  labs(title = "Monthly Shootings: Before vs After NFSU", x = "Month", y = "Count")
```

## Visualization 5.2 — Monthly Counts with NFSU Intervention Line

```{r}
ggplot(monthly_counts, aes(x = month, y = count)) +
  geom_line() +
  geom_vline(xintercept = as.Date("2024-01-01"), color = "red") +
  labs(title = "Monthly Shootings Over Time (NFSU Start Marked)", x = "Month", y = "Count")
```

## Visualization 5.3 — Yearly Counts (Bar)

```{r}
year_counts <- map_data %>%
  group_by(Year) %>%
  summarize(count = n(), .groups = "drop")

ggplot(year_counts, aes(x = Year, y = count)) +
  geom_col() +
  labs(title = "Shootings by Year", x = "Year", y = "Count")
```

```{r}
year_period_counts <- map_data %>%
  group_by(period, Year) %>%
  summarize(count = n(), .groups = "drop")

ggplot(year_period_counts, aes(x = Year, y = count)) +
  geom_col() +
  facet_wrap(~period, scales = "free_x") +
  labs(title = "Yearly Shootings: Before vs After NFSU", x = "Year", y = "Count")
```

## Visualization 5.4 — Before vs After NFSU (Bar)

```{r}
period_counts <- map_data %>%
  group_by(period) %>%
  summarize(count = n(), .groups = "drop")

ggplot(period_counts, aes(x = period, y = count)) +
  geom_col() +
  labs(title = "Shootings Before vs After NFSU", x = "", y = "Count")
```

## Visualization 5.5 — Time of Day Pattern (Histogram)

```{r}
ggplot(map_data, aes(x = hour)) +
  geom_bar() +
  labs(title = "Shootings by Hour of Day", x = "Hour", y = "Count")
```

```{r}
ggplot(map_data, aes(x = hour)) +
  geom_bar() +
  facet_wrap(~period) +
  labs(title = "Shootings by Hour of Day: Before vs After NFSU", x = "Hour", y = "Count")
```

## Visualization 5.6 — Day of Week Pattern (Bar)

```{r}
ggplot(map_data, aes(x = day)) +
  geom_bar() +
  labs(title = "Shootings by Day of Week", x = "Day", y = "Count")
```

```{r}
ggplot(map_data, aes(x = day)) +
  geom_bar() +
  facet_wrap(~period) +
  labs(title = "Shootings by Day of Week: Before vs After NFSU", x = "Day", y = "Count")
```

## Visualization 5.7 — Seasonal Patterns

```{r}
map_data <- map_data %>%
  mutate(
    season = case_when(
      month(DATE) %in% c(12, 1, 2) ~ "Winter",
      month(DATE) %in% c(3, 4, 5)  ~ "Spring",
      month(DATE) %in% c(6, 7, 8)  ~ "Summer",
      month(DATE) %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  mutate(
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    period = factor(period, levels = c("Before NFSU", "After NFSU"))
  )
```

```{r}
season_counts <- map_data %>%
  group_by(period, season) %>%
  summarize(count = n(), .groups = "drop")

ggplot(season_counts, aes(x = season, y = count)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~period) +
  labs(
    title    = "Shootings by Season",
    subtitle = "Before vs After NFSU",
    x        = "Season",
    y        = "Count"
  ) +
  theme_minimal()
```

```{r}
ggplot(season_counts, aes(x = season, y = count, fill = period)) +
  geom_col(position = "dodge") +
  labs(title = "Seasonal Shootings Comparison", x = "Season", y = "Count") +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 3: Spatial Patterns

## Visualization 5.8 — Spatial Density (Heat Map)

```{r}
ggplot(map_data, aes(x = Long, y = Lat)) +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon") +
  labs(title = "Shooting Density Map", x = "Longitude", y = "Latitude")
```

```{r}
ggplot(map_data, aes(x = Long, y = Lat)) +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon") +
  facet_wrap(~period) +
  labs(title = "Shooting Density Map: Before vs After NFSU", x = "Longitude", y = "Latitude")
```

------------------------------------------------------------------------

# Section 4: Time Between Events

## Prepare Time-Gap Data

```{r}
shootings <- read_excel("Shootings_Full Data(new).xlsx", sheet = "Full Data")

shootings_gaps <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    )
  ) %>%
  filter(!is.na(event_dt)) %>%
  arrange(event_dt) %>%
  mutate(
    prev_event_dt = lag(event_dt),
    gap_minutes   = as.numeric(difftime(event_dt, prev_event_dt, units = "mins")),
    gap_hours     = as.numeric(difftime(event_dt, prev_event_dt, units = "hours")),
    gap_days      = as.numeric(difftime(event_dt, prev_event_dt, units = "days"))
  ) %>%
  filter(!is.na(gap_hours))

nrow(shootings_gaps)
head(shootings_gaps)
```

## Visualization 6.1 — Time Between Events: Hours (Histogram)

```{r}
ggplot(shootings_gaps %>% filter(gap_hours <= 168), aes(x = gap_hours)) +
  geom_histogram(binwidth = 4) +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Time Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Hours Since Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Visualization 6.2 — Time Between Events: First 24 Hours (Count)

```{r}
ggplot(shootings_gaps %>% filter(gap_minutes <= 1440), aes(x = gap_minutes)) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Time Between Consecutive Shooting Events",
    subtitle = "First 24 Hours | Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Visualization 6.3 — Time Between Events: First 24 Hours (Normalized)

```{r}
ggplot(shootings_gaps %>% filter(gap_minutes <= 1440),
       aes(x = gap_minutes, y = after_stat(density))) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distribution of Time Between Events",
    subtitle = "Normalized Comparison (First 24 Hours)",
    x        = "Minutes Since Previous Event",
    y        = "Density"
  ) +
  theme_minimal()
```

## Visualization 6.4 — Time Between Events: Overlay

```{r}
ggplot(shootings_gaps %>% filter(gap_minutes <= 1440),
       aes(x = gap_minutes, fill = period)) +
  geom_histogram(binwidth = 30, alpha = 0.5, position = "identity") +
  labs(
    title    = "Time Between Events (Overlay)",
    subtitle = "Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Data Quality Check

```{r}
summary(shootings$Time)
sum(is.na(shootings$Time))
sum(is.na(shootings_gaps$event_dt))
table(shootings_gaps$period)
nrow(shootings_gaps)
```

------------------------------------------------------------------------

# Section 5: Distance Between Events

## Prepare Distance Data

```{r}
shootings_space <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    )
  ) %>%
  filter(!is.na(event_dt), !is.na(Lat), !is.na(Long)) %>%
  arrange(event_dt) %>%
  mutate(
    prev_lat   = lag(Lat),
    prev_long  = lag(Long),
    dist_meters = distHaversine(
      matrix(c(prev_long, prev_lat), ncol = 2),
      matrix(c(Long, Lat),           ncol = 2)
    ),
    dist_miles = dist_meters / 1609.34,
    dist_feet  = dist_meters * 3.28084
  ) %>%
  filter(!is.na(dist_miles))
```

## Visualization 7.1 — Distance Between Events: Miles (Count)

```{r}
ggplot(shootings_space %>% filter(dist_miles <= 5), aes(x = dist_miles)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distance Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Miles from Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Visualization 7.2 — Distance Between Events: Miles (Normalized)

```{r}
ggplot(shootings_space %>% filter(dist_miles <= 5),
       aes(x = dist_miles, y = after_stat(density))) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distribution of Distance Between Events",
    subtitle = "Normalized Comparison | Before vs After NFSU",
    x        = "Miles from Previous Event",
    y        = "Density"
  ) +
  theme_minimal()
```

## Visualization 7.3 — Distance Between Events: Feet

```{r}
ggplot(shootings_space %>% filter(dist_feet <= 10000), aes(x = dist_feet)) +
  geom_histogram(binwidth = 500, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distance Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Feet from Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Summary Statistics — Distance

```{r}
shootings_space %>%
  group_by(period) %>%
  summarize(
    n                 = n(),
    mean_dist_miles   = mean(dist_miles,   na.rm = TRUE),
    median_dist_miles = median(dist_miles, na.rm = TRUE),
    sd_dist_miles     = sd(dist_miles,     na.rm = TRUE),
    min_dist_miles    = min(dist_miles,    na.rm = TRUE),
    max_dist_miles    = max(dist_miles,    na.rm = TRUE)
  )
```

------------------------------------------------------------------------

# Section 6: Space-Time Heatmaps

## Prepare Day × Hour Heatmap Data

```{r}
shootings_time <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = if_else(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
    hour     = hour(event_dt),
    day      = wday(event_dt, label = TRUE, week_start = 1)
  ) %>%
  filter(!is.na(event_dt)) %>%
  mutate(
    day    = factor(day,    levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    period = factor(period, levels = c("Before NFSU", "After NFSU"))
  )

heatmap_data <- shootings_time %>%
  group_by(period, day, hour) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(
    period = factor(period, levels = c("Before NFSU", "After NFSU")),
    day    = factor(day,    levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
  )
```

## Visualization 8.1 — Day × Hour Heatmap (Raw Counts)

```{r}
ggplot(heatmap_data, aes(x = hour, y = day, fill = count)) +
  geom_tile() +
  facet_wrap(~period) +
  scale_fill_viridis_c() +
  labs(
    title    = "Shootings by Day of Week and Hour",
    subtitle = "Before vs After NFSU",
    x        = "Hour of Day",
    y        = "Day of Week",
    fill     = "Count"
  ) +
  theme_minimal()
```

## Visualization 8.2 — Day × Hour Heatmap (Normalized)

```{r}
heatmap_data_norm <- heatmap_data %>%
  group_by(period) %>%
  mutate(prop = count / sum(count))

ggplot(heatmap_data_norm, aes(x = hour, y = day, fill = prop)) +
  geom_tile() +
  facet_wrap(~period) +
  scale_fill_viridis_c() +
  labs(
    title    = "Distribution of Shootings by Day & Hour",
    subtitle = "Normalized | Before vs After NFSU",
    x        = "Hour of Day",
    y        = "Day of Week",
    fill     = "Proportion"
  ) +
  theme_minimal()
```

## Visualization 8.3 — Hourly Patterns by Day of Week (Line)

```{r}
heatmap_data %>%
  ggplot(aes(x = hour, y = count, color = day)) +
  geom_line() +
  facet_wrap(~period) +
  labs(
    title = "Hourly Shooting Patterns by Day of Week",
    x     = "Hour of Day",
    y     = "Count"
  ) +
  theme_dark()
```

## Visualization 8.4 — Hourly Patterns by Day (Facet Grid)

```{r}
heatmap_data %>%
  ggplot(aes(x = hour, y = count)) +
  geom_line(color = "steelblue") +
  facet_grid(day ~ period) +
  labs(
    title = "Hourly Shooting Patterns by Day",
    x     = "Hour of Day",
    y     = "Count"
  )
```

## Visualization 8.5 — Change in Shootings After NFSU (Difference Heatmap)

```{r}
diff_data <- heatmap_data %>%
  tidyr::pivot_wider(names_from = period, values_from = count, values_fill = 0) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)

ggplot(diff_data, aes(x = hour, y = day, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(
    title = "Change in Shootings (After - Before NFSU)",
    x     = "Hour",
    y     = "Day",
    fill  = "Change"
  ) +
  theme_minimal()
```

## Visualization 8.6 — Polar Bar Chart (Hour × Day)

```{r}
heatmap_data %>%
  ggplot(aes(x = hour, y = count, fill = day)) +
  geom_bar(stat = "identity") +
  coord_polar() +
  facet_wrap(~period) +
  theme_minimal()
```

## Visualization 8.7 — Hourly Counts by Period (Grouped Bar)

```{r}
shootings_time %>%
  ggplot(aes(x = factor(hour), fill = period)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Shootings by Hour",
    x     = "Hour of Day",
    y     = "Count"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 7: Joint Space-Time Analysis

## Prepare Space-Time Data

```{r}
space_time <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = if_else(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU")
  ) %>%
  filter(!is.na(event_dt), !is.na(Lat), !is.na(Long)) %>%
  arrange(event_dt) %>%
  mutate(
    prev_event_dt = lag(event_dt),
    prev_lat      = lag(Lat),
    prev_long     = lag(Long),
    gap_minutes   = as.numeric(difftime(event_dt, prev_event_dt, units = "mins")),
    dist_meters   = distHaversine(
      matrix(c(prev_long, prev_lat), ncol = 2),
      matrix(c(Long, Lat),           ncol = 2)
    ),
    dist_miles = dist_meters / 1609.34
  ) %>%
  filter(!is.na(gap_minutes), !is.na(dist_miles)) %>%
  mutate(period = factor(period, levels = c("Before NFSU", "After NFSU")))
```

## Visualization 9.1 — Time vs Distance (2D Bin)

```{r}
ggplot(space_time %>% filter(gap_minutes <= 1440, dist_miles <= 5),
       aes(x = gap_minutes, y = dist_miles)) +
  geom_bin2d() +
  facet_wrap(~period) +
  labs(
    title    = "Time and Distance Between Consecutive Shooting Events",
    subtitle = "First 24 Hours and Within 5 Miles | Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Miles From Previous Event",
    fill     = "Count"
  ) +
  theme_minimal()
```

## Visualization 9.2 — Time vs Distance (Scatter)

```{r}
ggplot(space_time %>% filter(gap_minutes <= 1440, dist_miles <= 5),
       aes(x = gap_minutes, y = dist_miles)) +
  geom_point(alpha = 0.25, size = 1) +
  facet_wrap(~period) +
  labs(
    title    = "Time and Distance Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Miles From Previous Event"
  ) +
  theme_minimal()
```

## Visualization 9.3 — Time vs Distance (Joint Density)

```{r}
ggplot(space_time %>% filter(gap_minutes <= 1440, dist_miles <= 5),
       aes(x = gap_minutes, y = dist_miles)) +
  geom_density_2d_filled(contour_var = "ndensity") +
  facet_wrap(~period) +
  labs(
    title    = "Joint Distribution of Time and Distance Between Events",
    subtitle = "Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Miles From Previous Event",
    fill     = "Density"
  ) +
  theme_minimal()
```
