---
title: "DASC 460: M2a vs Neighborhood-Cluster Background Model"
subtitle: "Frogtown · Phalen (High-High) + Mac-Groveland/Highland (Low-Low) vs Global Background"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
editor_options:
  markdown:
    wrap: 72
---

------------------------------------------------------------------------

# Setup

```{r setup, warning=FALSE, message=FALSE}
library(rmarkdown); library(knitr)
library(readxl);    library(dplyr)
library(ggplot2);   library(lubridate)
library(tidyr);     library(sf)
library(xml2);      library(spdep)
library(surveillance); library(patchwork)
```

------------------------------------------------------------------------

# Section 1: Load Data & Build Panel

## 1.1 Shooting Records

```{r load-data}
file <- "Shootings_Full Data(new).xlsx"

full_data <- read_excel(file, sheet = "Full Data") %>%
  mutate(
    DATE       = as.Date(DATE),
    week       = floor_date(DATE, "week", week_start = 1),
    week_index = as.integer(difftime(week, min(week, na.rm = TRUE), units = "weeks")),
    nfsu       = as.integer(DATE >= as.Date("2024-01-01")),
    period     = factor(if_else(DATE < as.Date("2024-01-01"),
                                "Before NFSU", "After NFSU"),
                        levels = c("Before NFSU", "After NFSU")),
    season     = factor(case_when(
      month(DATE) %in% c(12, 1, 2)  ~ "Winter",
      month(DATE) %in% c(3, 4, 5)   ~ "Spring",
      month(DATE) %in% c(6, 7, 8)   ~ "Summer",
      month(DATE) %in% c(9, 10, 11) ~ "Fall"
    ), levels = c("Winter", "Spring", "Summer", "Fall")),
    GRID = as.character(GRID)
  ) %>%
  filter(!is.na(Lat), !is.na(Long), !is.na(DATE), Lat != 0, Long != 0)

cat("Shooting records loaded:", nrow(full_data), "\n")
```

## 1.2 Police Grid (KMZ)

```{r load-grid, message=FALSE, warning=FALSE}
kmz_path <- "StPaul_Police_Gri_LayerToKML.kmz"
unzip(kmz_path, exdir = tempdir(), overwrite = TRUE)
kml_path <- list.files(tempdir(), pattern = "\\.kml$", full.names = TRUE)[1]

kml_doc    <- read_xml(kml_path)
ns         <- c(kml = "http://www.opengis.net/kml/2.2")
placemarks <- xml_find_all(kml_doc, ".//kml:Placemark", ns)

features <- lapply(placemarks, function(pm) {
  grid_id    <- xml_text(xml_find_first(pm, "kml:name", ns))
  poly_nodes <- xml_find_all(pm,
    "kml:MultiGeometry/kml:Polygon/kml:outerBoundaryIs/kml:LinearRing/kml:coordinates", ns)
  if (length(poly_nodes) == 0)
    poly_nodes <- xml_find_all(pm,
      "kml:Polygon/kml:outerBoundaryIs/kml:LinearRing/kml:coordinates", ns)

  coords_list <- lapply(poly_nodes, function(co) {
    pts <- strsplit(trimws(xml_text(co)), "\\s+")[[1]]
    do.call(rbind, lapply(pts, function(p) {
      xy <- as.numeric(strsplit(p, ",")[[1]]); c(xy[1], xy[2])
    }))
  })

  geom <- if (length(coords_list) == 1) st_polygon(list(coords_list[[1]]))
          else st_multipolygon(lapply(coords_list, function(m) list(m)))
  st_sf(Name = grid_id, geometry = st_sfc(geom, crs = 4326))
})

grid_sf <- do.call(rbind, features) %>% mutate(Name = as.character(Name))
cat("Grid cells loaded:", nrow(grid_sf), "\n")
```

## 1.3 Spatial Join & Weekly Panel

```{r build-panel}
shootings_sf <- full_data %>%
  st_as_sf(coords = c("Long", "Lat"), crs = 4326)

shootings_gridded <- st_join(shootings_sf, grid_sf, join = st_within) %>%
  rename(grid_id = Name) %>%
  st_drop_geometry()

cat("Matched:", sum(!is.na(shootings_gridded$grid_id)),
    "of", nrow(shootings_gridded), "records to a grid\n")

all_weeks <- sort(unique(shootings_gridded$week))
all_grids <- sort(unique(grid_sf$Name))

panel <- expand.grid(week = all_weeks, grid_id = all_grids,
                     stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  mutate(week = as.Date(week)) %>%
  left_join(
    shootings_gridded %>%
      filter(!is.na(grid_id), !is.na(week)) %>%
      group_by(week, grid_id) %>%
      summarize(count = n(), .groups = "drop"),
    by = c("week", "grid_id")
  ) %>%
  replace_na(list(count = 0)) %>%
  mutate(
    nfsu       = as.integer(week >= as.Date("2024-01-01")),
    season     = factor(case_when(
      month(week) %in% c(12, 1, 2)  ~ "Winter",
      month(week) %in% c(3, 4, 5)   ~ "Spring",
      month(week) %in% c(6, 7, 8)   ~ "Summer",
      month(week) %in% c(9, 10, 11) ~ "Fall"
    ), levels = c("Winter", "Spring", "Summer", "Fall")),
    week_index = as.integer(difftime(week, min(week), units = "weeks"))
  )

cat("Panel:", nrow(panel), "rows |",
    length(all_weeks), "weeks ×", length(all_grids), "grids\n")
cat("Overall zero proportion:", round(mean(panel$count == 0), 3), "\n")
```

## 1.4 sts Object & Shared Matrices

```{r build-sts}
all_grid_ids  <- sort(unique(panel$grid_id))
all_week_vals <- sort(unique(panel$week))

count_wide <- panel %>%
  dplyr::select(week, grid_id, count) %>%
  pivot_wider(names_from = grid_id, values_from = count, values_fill = 0) %>%
  arrange(week)

count_matrix <- count_wide %>% dplyr::select(-week) %>% as.matrix()
rownames(count_matrix) <- as.character(count_wide$week)

grid_sf_ordered <- grid_sf %>%
  rename(grid_id = Name) %>%
  filter(grid_id %in% colnames(count_matrix)) %>%
  arrange(match(grid_id, colnames(count_matrix)))

nb_full <- poly2nb(grid_sf_ordered, queen = TRUE)
adj_mat <- nb2mat(nb_full, style = "B", zero.policy = TRUE)
rownames(adj_mat) <- colnames(adj_mat) <- grid_sf_ordered$grid_id

sts_obj <- sts(observed = count_matrix, neighbourhood = adj_mat, frequency = 52)

nfsu_vec    <- as.integer(all_week_vals >= as.Date("2024-01-01"))
nfsu_matrix <- matrix(rep(nfsu_vec, ncol(count_matrix)),
                      nrow = nrow(count_matrix), ncol = ncol(count_matrix))
n_obs <- sum(count_matrix)
grid_order <- colnames(count_matrix)

cat("sts object:", nrow(sts_obj), "weeks ×", ncol(sts_obj), "grids\n")
cat("Before NFSU:", sum(nfsu_vec == 0), "weeks |",
    "After NFSU:", sum(nfsu_vec == 1), "weeks\n")
```

------------------------------------------------------------------------

# Section 2: Cluster Identification

## 2.1 Global Moran's I

Confirms significant spatial autocorrelation — a prerequisite for
meaningful LISA.

```{r morans-i}
grid_totals <- panel %>%
  group_by(grid_id) %>%
  summarize(total = sum(count), .groups = "drop")

nb_lisa <- poly2nb(grid_sf_ordered, queen = TRUE)
w_lisa  <- nb2listw(nb_lisa, style = "W", zero.policy = TRUE)

totals_ordered <- grid_totals$total[
  match(grid_sf_ordered$grid_id, grid_totals$grid_id)]
totals_ordered[is.na(totals_ordered)] <- 0

moran_result <- moran.test(totals_ordered, w_lisa, zero.policy = TRUE)
cat("=== Global Moran's I (Full Period) ===\n")
cat(sprintf("I = %.4f   p-value = %s\n",
    moran_result$estimate["Moran I statistic"],
    format(moran_result$p.value, digits = 4)))
cat("Significant spatial autocorrelation:", moran_result$p.value < 0.05, "\n")
```

## 2.2 LISA — Full Period, Before & After NFSU

```{r run-lisa}
run_lisa <- function(counts_vec, w, label) {
  set.seed(42)
  res      <- localmoran_perm(counts_vec, w, nsim = 999,
                              zero.policy = TRUE, alternative = "two.sided")
  lag_vals <- lag.listw(w, counts_vec, zero.policy = TRUE)
  tibble(
    grid_id   = grid_sf_ordered$grid_id,
    total     = counts_vec,
    li        = res[, "Ii"],
    z_score   = res[, "Z.Ii"],
    p_value   = res[, "Pr(z != E(Ii))"],
    lag_total = lag_vals,
    period    = label
  ) %>%
    mutate(
      cluster_type = factor(case_when(
        p_value < 0.05 & total    > mean(total) & lag_total > mean(total) ~ "High-High",
        p_value < 0.05 & total    < mean(total) & lag_total < mean(total) ~ "Low-Low",
        p_value < 0.05 & total    > mean(total) & lag_total < mean(total) ~ "High-Low",
        p_value < 0.05 & total    < mean(total) & lag_total > mean(total) ~ "Low-High",
        TRUE ~ "Not significant"
      ), levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not significant"))
    )
}

get_ordered_totals <- function(df) {
  v <- df$total[match(grid_sf_ordered$grid_id, df$grid_id)]
  v[is.na(v)] <- 0; v
}

before_totals <- panel %>% filter(nfsu == 0) %>%
  group_by(grid_id) %>% summarize(total = sum(count), .groups = "drop")
after_totals  <- panel %>% filter(nfsu == 1) %>%
  group_by(grid_id) %>% summarize(total = sum(count), .groups = "drop")

lisa_full   <- run_lisa(totals_ordered,                    w_lisa, "Full period")
lisa_before <- run_lisa(get_ordered_totals(before_totals), w_lisa, "Before NFSU")
lisa_after  <- run_lisa(get_ordered_totals(after_totals),  w_lisa, "After NFSU")

cat("=== LISA Summary — Full Period ===\n")
print(lisa_full %>% count(cluster_type))
```

### Grids That Changed Status (Before → After NFSU)

```{r changed-grids}
changed_grids <- lisa_before %>%
  dplyr::select(grid_id, before_cluster = cluster_type) %>%
  left_join(lisa_after %>% dplyr::select(grid_id, after_cluster = cluster_type),
            by = "grid_id") %>%
  filter(as.character(before_cluster) != as.character(after_cluster)) %>%
  arrange(before_cluster, after_cluster)

cat("Grids that changed cluster status:", nrow(changed_grids), "\n\n")
print(changed_grids)
```

------------------------------------------------------------------------

# Section 3: Cluster Assignment

Based on the LISA results and the St. Paul Police grid map, the 27
High-High grids split into two geographically distinct hot-spot
neighborhoods. The 5 Low-Low grids form a single cold-spot cluster in
the southwest.

```{r cluster-assign}
# ── CLUSTER GRID ASSIGNMENTS ──────────────────────────────────────────────────
# Frogtown (Thomas-Dale): Dale/University/Western corridor — central St. Paul
frogtown_grids    <- c("30", "50", "67", "68", "70", "87", "88", "89", "90", "107", "108", "109", "110")

# Phalen: Arcade/Payne/Johnson Pkwy — northeast St. Paul
phalen_grids      <- c("33", "34", "54", "55", "57", "73", "74", "75", "76",
                       "94", "95", "96", "114", "116")

# Mac-Groveland / Highland Park — persistent Low-Low cold spot, southwest
mac_highland_grids <- c("164", "184", "185", "204", "205")
# ──────────────────────────────────────────────────────────────────────────────

cluster_list <- list(
  frogtown     = frogtown_grids,
  phalen       = phalen_grids,
  mac_highland = mac_highland_grids
)

all_cluster_grids <- unique(unlist(cluster_list))

cat("=== Cluster Assignment ===\n\n")
for (nm in names(cluster_list)) {
  cat(sprintf("  %-15s  %2d grids: %s\n",
      nm, length(cluster_list[[nm]]),
      paste(sort(cluster_list[[nm]]), collapse = ", ")))
}
cat(sprintf("\n  %-15s  %2d grids (reference)\n",
    "Other", ncol(count_matrix) - length(all_cluster_grids)))
```

------------------------------------------------------------------------

# Section 4: Background Rate per Cluster

How often does each cluster get shot at, on average, per grid per week?
This is the empirical baseline the model's endemic component must
capture.

```{r background-rates}
# Assign cluster label to every grid
grid_cluster_map <- tibble(grid_id = grid_order) %>%
  mutate(cluster = case_when(
    grid_id %in% frogtown_grids     ~ "Frogtown",
    grid_id %in% phalen_grids       ~ "Phalen",
    grid_id %in% mac_highland_grids ~ "Mac-Highland",
    TRUE                            ~ "Other (reference)"
  ))

grid_weekly_mean <- panel %>%
  group_by(grid_id) %>%
  summarize(mean_weekly = mean(count),
            sd_weekly   = sd(count),
            zero_prop   = mean(count == 0),
            .groups = "drop") %>%
  left_join(grid_cluster_map, by = "grid_id")

# Summary table by cluster
cluster_summary <- grid_weekly_mean %>%
  group_by(cluster) %>%
  summarize(
    n_grids          = n(),
    mean_rate        = mean(mean_weekly),
    sd_rate          = sd(mean_weekly),
    median_zero_prop = median(zero_prop),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_rate))

cat("=== Mean Weekly Shooting Rate & Sparsity by Cluster ===\n\n")
print(cluster_summary, n = Inf)
```

## 4.1 Background Rate Bar Chart

```{r bg-rate-plot, fig.height=3.5, fig.width=7}
cluster_colors <- c(
  "Frogtown"          = "#d7191c",
  "Phalen"            = "#e85d1a",
  "Mac-Highland"      = "#2c7bb6",
  "Other (reference)" = "#aaaaaa"
)

ggplot(cluster_summary,
       aes(x = reorder(cluster, mean_rate), y = mean_rate, fill = cluster)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_errorbar(aes(ymin = pmax(mean_rate - sd_rate, 0),
                    ymax = mean_rate + sd_rate),
                width = 0.2, color = "grey30") +
  coord_flip() +
  scale_fill_manual(values = cluster_colors) +
  labs(
    title    = "Empirical Background Rate by Cluster",
    subtitle = "Mean ± SD of weekly shooting count per grid",
    x = NULL, y = "Mean weekly count per grid"
  ) +
  theme_minimal(base_size = 12)
```

## 4.2 Zero-Count Sparsity by Cluster

```{r sparsity-plot, fig.height=4, fig.width=9}
# Per-grid zero proportions, ordered within cluster
grid_weekly_mean <- grid_weekly_mean %>%
  mutate(cluster = factor(cluster,
    levels = c("Frogtown", "Phalen", "Mac-Highland", "Other (reference)")))

p_strip <- ggplot(grid_weekly_mean,
                  aes(x = reorder(grid_id, zero_prop), y = zero_prop,
                      fill = cluster)) +
  geom_col(width = 0.8, show.legend = FALSE) +
  facet_wrap(~cluster, scales = "free_x", nrow = 1) +
  scale_fill_manual(values = cluster_colors) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title    = "Zero-Count Proportion per Grid, by Cluster",
    subtitle = "Each bar = one grid | Higher = more sparse",
    x = "Grid (ordered by sparsity within cluster)", y = "% weeks with zero shootings"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_box <- ggplot(grid_weekly_mean,
                aes(x = cluster, y = zero_prop, fill = cluster)) +
  geom_boxplot(width = 0.5, outlier.size = 0.8, show.legend = FALSE) +
  scale_fill_manual(values = cluster_colors) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Sparsity Distribution by Cluster",
       x = NULL, y = "% weeks with zero shootings") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

p_strip / p_box + plot_layout(heights = c(2, 1))
```

```{r sparsity-summary}
cat("=== Sparsity Summary by Cluster ===\n\n")
cat(sprintf("Overall panel zero proportion: %.1f%%\n\n",
    mean(panel$count == 0) * 100))

grid_weekly_mean %>%
  group_by(cluster) %>%
  summarize(
    grids        = n(),
    median_zeros = scales::percent(median(zero_prop), accuracy = 0.1),
    min_zeros    = scales::percent(min(zero_prop),    accuracy = 0.1),
    max_zeros    = scales::percent(max(zero_prop),    accuracy = 0.1),
    .groups = "drop"
  ) %>%
  arrange(median_zeros) %>%
  print(n = Inf)
```

------------------------------------------------------------------------

# Section 5: Spatial Visualizations

## 5.1 Shooting Heatmap (Total Counts per Grid)

```{r heatmap, fig.height=5.5, fig.width=6}
heatmap_sf <- grid_sf_ordered %>%
  left_join(grid_totals, by = "grid_id")

ggplot(heatmap_sf) +
  geom_sf(aes(fill = total), color = "white", linewidth = 0.15) +
  scale_fill_gradientn(
    colours  = c("#f7f7f7", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f"),
    name     = "Total\nshootings",
    na.value = "grey90"
  ) +
  labs(
    title    = "Total Non-Fatal Shootings per Grid",
    subtitle = "Full study period (all weeks)"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## 5.2 LISA Cluster Maps

```{r lisa-maps, fig.height=5, fig.width=11}
lisa_color <- c(
  "High-High"       = "#d7191c",
  "Low-Low"         = "#2c7bb6",
  "High-Low"        = "#fdae61",
  "Low-High"        = "#abd9e9",
  "Not significant" = "#f0f0f0"
)

plot_lisa_map <- function(lisa_df, label) {
  grid_sf_ordered %>%
    left_join(lisa_df %>% dplyr::select(grid_id, cluster_type), by = "grid_id") %>%
    ggplot() +
    geom_sf(aes(fill = cluster_type), color = "white", linewidth = 0.2) +
    scale_fill_manual(values = lisa_color, name = "LISA cluster\n(p < 0.05)",
                      drop = FALSE) +
    labs(title = label) +
    theme_minimal() +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          legend.position = "bottom", plot.title = element_text(size = 11))
}

plot_lisa_map(lisa_before, "Before NFSU") +
plot_lisa_map(lisa_after,  "After NFSU")
```

```{r lisa-full-map, fig.height=5.5, fig.width=6}
plot_lisa_map(lisa_full, "Full Study Period") +
  theme(legend.position = "right")
```

## 5.3 Neighborhood Cluster Assignment Map

```{r cluster-map, fig.height=5.5, fig.width=6}
neighborhood_colors <- c(
  "Frogtown"          = "#d7191c",
  "Phalen"            = "#e85d1a",
  "Mac-Highland"      = "#2c7bb6",
  "Other (reference)" = "#f0f0f0"
)

grid_sf_ordered %>%
  left_join(grid_cluster_map, by = "grid_id") %>%
  mutate(cluster = factor(cluster,
    levels = c("Frogtown", "Phalen", "Mac-Highland", "Other (reference)"))) %>%
  ggplot() +
  geom_sf(aes(fill = cluster), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = neighborhood_colors, name = "Neighborhood\ncluster") +
  labs(
    title    = "Neighborhood Cluster Assignment Map",
    subtitle = "Two High-High hot spots · One Low-Low cold spot · Reference"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        legend.position = "right")
```

------------------------------------------------------------------------

# Section 6: Build Cluster Covariate Matrices

```{r cluster-matrices}
make_cluster_matrix <- function(ids) {
  flag <- as.integer(grid_order %in% ids)
  matrix(flag, nrow = nrow(count_matrix), ncol = ncol(count_matrix),
         byrow = TRUE, dimnames = list(NULL, grid_order))
}

mat_frogtown     <- make_cluster_matrix(frogtown_grids)
mat_phalen       <- make_cluster_matrix(phalen_grids)
mat_mac_highland <- make_cluster_matrix(mac_highland_grids)

# "Other" = all grids not in any named cluster — this is the endemic reference group
other_grids  <- setdiff(grid_order, all_cluster_grids)
mat_other    <- make_cluster_matrix(other_grids)

cat("Covariate matrices built:\n")
cat(sprintf("  frogtown     : %d grids flagged\n", sum(mat_frogtown[1, ])))
cat(sprintf("  phalen       : %d grids flagged\n", sum(mat_phalen[1, ])))
cat(sprintf("  mac_highland : %d grids flagged\n", sum(mat_mac_highland[1, ])))
cat(sprintf("  other (ref)  : %d grids (reference — captured by end.1 intercept)\n",
    sum(mat_other[1, ])))
cat(sprintf("  TOTAL        : %d grids (should equal %d)\n",
    sum(mat_frogtown[1,]) + sum(mat_phalen[1,]) +
    sum(mat_mac_highland[1,]) + sum(mat_other[1,]),
    ncol(count_matrix)))
```

------------------------------------------------------------------------

# Section 7: Model Fitting

## 7.1 Baseline — M1 (Warm Start)

```{r fit-m1}
ctrl_m1 <- list(ar = list(f = ~ 1), end = list(f = ~ 1), family = "Poisson")
fit_m1  <- tryCatch(hhh4(sts_obj, control = ctrl_m1),
                    error = function(e) { cat("M1:", conditionMessage(e)); NULL })
cat("M1 converged:", !is.null(fit_m1) && fit_m1$convergence, "\n")
```

## 7.2 Model M2a — Global Background + NFSU-Stratified φ

$$\lambda_{i,t} = e^{\beta_0} + \phi_j \cdot Y_{i,t-1}$$

One background rate for all grids. $\phi_j$ differs before vs. after
NFSU. This is the **baseline** for comparison.

```{r fit-m2a}
ctrl_m2a <- list(
  ar        = list(f = ~ 1 + nfsu),
  end       = list(f = ~ 1),
  family    = "Poisson",
  data      = list(nfsu = nfsu_matrix),
  optimizer = list(stop = list(iter.max = 300, tol = 1e-6)),
  start     = list(fixed = c(
    "ar.1"    = coef(fit_m1)["ar.1"],
    "ar.nfsu" = 0,
    "end.1"   = coef(fit_m1)["end.1"]
  ))
)

fit_m2a <- suppressWarnings(tryCatch(
  hhh4(sts_obj, control = ctrl_m2a),
  error = function(e) { cat("M2a:", conditionMessage(e)); NULL }
))

if (!is.null(fit_m2a)) {
  cat("=== M2a: Global Background ===\n")
  cat("Converged:", fit_m2a$convergence, "\n\n")
  print(summary(fit_m2a))

  c0  <- fit_m2a$coefficients
  se0 <- sqrt(diag(fit_m2a$cov))

  # φ before NFSU: exp(ar.1)
  log_phi_before <- c0["ar.1"]
  se_phi_before  <- se0["ar.1"]

  # φ after NFSU: exp(ar.1 + ar.nfsu)  — delta-method SE
  log_phi_after  <- c0["ar.1"] + c0["ar.nfsu"]
  se_phi_after   <- sqrt(se0["ar.1"]^2 + se0["ar.nfsu"]^2 +
                           2 * fit_m2a$cov["ar.1", "ar.nfsu"])

  cat("\n--- φ (contagion) estimates with 95% CI (delta method) ---\n")
  cat(sprintf("  φ before NFSU : %.4f  [%.4f, %.4f]\n",
      exp(log_phi_before),
      exp(log_phi_before - 1.96 * se_phi_before),
      exp(log_phi_before + 1.96 * se_phi_before)))
  cat(sprintf("  φ after  NFSU : %.4f  [%.4f, %.4f]\n",
      exp(log_phi_after),
      exp(log_phi_after  - 1.96 * se_phi_after),
      exp(log_phi_after  + 1.96 * se_phi_after)))
  cat(sprintf("  Change        : %+.1f%%\n",
      (exp(log_phi_after) - exp(log_phi_before)) / exp(log_phi_before) * 100))
  cat(sprintf("\n  AIC: %.2f   BIC: %.2f\n",
      -2 * fit_m2a$loglikelihood + 2 * length(c0),
      -2 * fit_m2a$loglikelihood + log(n_obs) * length(c0)))
}
```

## 7.3 Model M2a(N) — Neighborhood Cluster Background + NFSU-Stratified φ

$$\lambda_{i,t} = e^{\,\beta_0 \;+\; \beta_{\text{FT}}\,\mathbf{1}[i \in \text{Frogtown}] \;+\; \beta_{\text{PH}}\,\mathbf{1}[i \in \text{Phalen}] \;+\; \beta_{\text{MH}}\,\mathbf{1}[i \in \text{Mac-Highland}]} + \phi_j \cdot Y_{i,t-1}$$

Three neighborhood-level background offsets; all other grids form the
reference ($\beta_0$). The AR structure is identical to M2a.

```{r fit-m2an}
ctrl_m2an <- list(
  ar        = list(f = ~ 1 + nfsu),
  end       = list(f = ~ 1 + frogtown + phalen + mac_highland),
  family    = "Poisson",
  data      = list(
    nfsu        = nfsu_matrix,
    frogtown    = mat_frogtown,
    phalen      = mat_phalen,
    mac_highland = mat_mac_highland
  ),
  optimizer = list(stop = list(iter.max = 500, tol = 1e-6)),
  start     = list(fixed = c(
    "ar.1"            = coef(fit_m2a)["ar.1"],
    "ar.nfsu"         = coef(fit_m2a)["ar.nfsu"],
    "end.1"           = coef(fit_m2a)["end.1"],
    "end.frogtown"    = 0,
    "end.phalen"      = 0,
    "end.mac_highland" = 0
  ))
)

fit_m2an <- suppressWarnings(tryCatch(
  hhh4(sts_obj, control = ctrl_m2an),
  error = function(e) { cat("M2a(N):", conditionMessage(e)); NULL }
))

if (!is.null(fit_m2an)) {
  cat("=== M2a(N): Neighborhood Cluster Background ===\n")
  cat("Converged:", fit_m2an$convergence, "\n\n")
  print(summary(fit_m2an))

  cn  <- fit_m2an$coefficients
  sen <- sqrt(diag(fit_m2an$cov))

  # φ before NFSU
  log_phi_before_n <- cn["ar.1"]
  se_phi_before_n  <- sen["ar.1"]

  # φ after NFSU — delta-method SE
  log_phi_after_n  <- cn["ar.1"] + cn["ar.nfsu"]
  se_phi_after_n   <- sqrt(sen["ar.1"]^2 + sen["ar.nfsu"]^2 +
                             2 * fit_m2an$cov["ar.1", "ar.nfsu"])

  cat("\n--- φ (contagion) estimates with 95% CI (delta method) ---\n")
  cat(sprintf("  φ before NFSU : %.4f  [%.4f, %.4f]\n",
      exp(log_phi_before_n),
      exp(log_phi_before_n - 1.96 * se_phi_before_n),
      exp(log_phi_before_n + 1.96 * se_phi_before_n)))
  cat(sprintf("  φ after  NFSU : %.4f  [%.4f, %.4f]\n",
      exp(log_phi_after_n),
      exp(log_phi_after_n  - 1.96 * se_phi_after_n),
      exp(log_phi_after_n  + 1.96 * se_phi_after_n)))
  cat(sprintf("  Change        : %+.1f%%\n",
      (exp(log_phi_after_n) - exp(log_phi_before_n)) / exp(log_phi_before_n) * 100))

  cat("\n--- Neighborhood background-rate offsets (reference = Other grids) ---\n")
  end_terms <- c("end.frogtown", "end.phalen", "end.mac_highland")
  for (nm in end_terms) {
    lo_rr <- exp(cn[nm] - 1.96 * sen[nm])
    hi_rr <- exp(cn[nm] + 1.96 * sen[nm])
    cat(sprintf("  %-25s  log(RR) = %+.4f   RR = %.3f  [%.3f, %.3f]\n",
        nm, cn[nm], exp(cn[nm]), lo_rr, hi_rr))
  }
  cat(sprintf("\n  AIC: %.2f   BIC: %.2f\n",
      -2 * fit_m2an$loglikelihood + 2 * length(cn),
      -2 * fit_m2an$loglikelihood + log(n_obs) * length(cn)))
}
```

------------------------------------------------------------------------

# Section 8: Model Comparison

## 8.1 φ Comparison — Before & After NFSU

```{r phi-comparison, fig.height=3.5, fig.width=7}
if (!is.null(fit_m2a) && !is.null(fit_m2an)) {
  c0 <- fit_m2a$coefficients
  cn <- fit_m2an$coefficients
  se0 <- sqrt(diag(fit_m2a$cov))
  sen <- sqrt(diag(fit_m2an$cov))

  phi_df <- tibble(
    model  = rep(c("M2a", "M2a(N)"), each = 2),
    period = rep(c("Before NFSU", "After NFSU"), 2),
    log_phi = c(c0["ar.1"],
                c0["ar.1"] + c0["ar.nfsu"],
                cn["ar.1"],
                cn["ar.1"] + cn["ar.nfsu"]),
    # Delta method SE for after: sqrt(var(ar.1) + var(ar.nfsu) + 2*cov)
    se_log = c(
      se0["ar.1"],
      sqrt(se0["ar.1"]^2 + se0["ar.nfsu"]^2 +
             2 * fit_m2a$cov["ar.1", "ar.nfsu"]),
      sen["ar.1"],
      sqrt(sen["ar.1"]^2 + sen["ar.nfsu"]^2 +
             2 * fit_m2an$cov["ar.1", "ar.nfsu"])
    )
  ) %>%
    mutate(
      phi = exp(log_phi),
      lo  = exp(log_phi - 1.96 * se_log),
      hi  = exp(log_phi + 1.96 * se_log),
      period = factor(period, levels = c("Before NFSU", "After NFSU"))
    )

  ggplot(phi_df, aes(x = period, y = phi, color = model, group = model)) +
    geom_line(linewidth = 0.8, linetype = "dashed") +
    geom_pointrange(aes(ymin = lo, ymax = hi),
                    position = position_dodge(width = 0.15), size = 0.7) +
    scale_color_manual(values = c("M2a" = "darkgreen", "M2a(N)" = "darkorange")) +
    labs(
      title    = "φ (Contagion Parameter) Before and After NFSU",
      subtitle = "Point estimates ± 95% CI (delta method) | Lower φ = less shooting-to-shooting spread",
      x = NULL, y = "φ", color = "Model"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

## 8.2 Neighborhood Background-Rate Ratios

```{r rr-plot, fig.height=3.5, fig.width=7}
if (!is.null(fit_m2an)) {
  cn  <- fit_m2an$coefficients
  sen <- sqrt(diag(fit_m2an$cov))
  end_nms <- c("end.frogtown", "end.phalen", "end.mac_highland")

  rr_df <- tibble(
    term    = end_nms,
    log_rr  = cn[end_nms],
    se      = sen[end_nms]
  ) %>%
    mutate(
      rr    = exp(log_rr),
      lo    = exp(log_rr - 1.96 * se),
      hi    = exp(log_rr + 1.96 * se),
      sig   = !(lo <= 1 & hi >= 1),
      label = c("Frogtown", "Phalen", "Mac-Highland")
    )

  ggplot(rr_df, aes(x = reorder(label, rr), y = rr, color = sig)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
    geom_pointrange(aes(ymin = lo, ymax = hi), size = 0.9) +
    coord_flip() +
    scale_color_manual(
      values = c("TRUE" = "#d7191c", "FALSE" = "grey60"),
      labels = c("TRUE" = "CI excludes 1", "FALSE" = "CI includes 1"),
      name = ""
    ) +
    labs(
      title    = "Neighborhood Background-Rate Ratios vs Reference",
      subtitle = "M2a(N) | Reference = non-cluster grids (Other)",
      x = NULL, y = "Rate Ratio exp(β)"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

## 8.3 AIC / BIC / Log-Likelihood Table

```{r fit-table}
make_row <- function(fit, name) {
  if (is.null(fit)) return(NULL)
  k  <- length(fit$coefficients)
  ll <- fit$loglikelihood
  tibble(Model = name, Params = k,
         LogLik = round(ll, 2),
         AIC    = round(-2 * ll + 2 * k, 2),
         BIC    = round(-2 * ll + log(n_obs) * k, 2),
         Converged = fit$convergence)
}

tbl <- bind_rows(
  make_row(fit_m2a,  "M2a  — Global background"),
  make_row(fit_m2an, "M2a(N) — Neighborhood clusters")
)
cat("=== Model Fit Statistics ===\n\n")
print(tbl, n = Inf)
```

## 8.4 Likelihood Ratio Test

```{r lrt}
if (!is.null(fit_m2a) && !is.null(fit_m2an)) {
  ll_0   <- fit_m2a$loglikelihood
  ll_1   <- fit_m2an$loglikelihood
  k_0    <- length(fit_m2a$coefficients)
  k_1    <- length(fit_m2an$coefficients)
  lrt    <- 2 * (ll_1 - ll_0)
  df_lrt <- k_1 - k_0
  p_lrt  <- pchisq(lrt, df = df_lrt, lower.tail = FALSE)

  cat(sprintf("LRT: M2a vs M2a(N)\n  χ² = %.4f   df = %d   p = %s\n\n",
      lrt, df_lrt, format(p_lrt, digits = 4)))
  if (p_lrt < 0.05) {
    cat("M2a(N) fits significantly better than M2a (p < 0.05).\n")
  } else {
    cat("No significant improvement over M2a (p ≥ 0.05).\n")
  }
}
```

------------------------------------------------------------------------

# Section 9: Diagnostics

Helper functions used throughout.

```{r diag-helpers}
get_fit_obs <- function(fit_obj) {
  fit_mat <- fitted(fit_obj)
  n_fit   <- nrow(fit_mat)
  idx     <- (nrow(count_matrix) - n_fit + 1):nrow(count_matrix)
  obs_mat <- observed(sts_obj)[idx, ]
  cols    <- intersect(colnames(obs_mat), colnames(fit_mat))
  list(fit = fit_mat[, cols, drop = FALSE],
       obs = obs_mat[, cols, drop = FALSE],
       wks = as.Date(rownames(count_matrix))[idx])
}

get_pearson_df <- function(fit_obj, label) {
  d   <- get_fit_obs(fit_obj)
  prs <- (d$obs - d$fit) / sqrt(d$fit + 1e-8)
  data.frame(resid = as.vector(prs), model = label)
}

get_deviance_df <- function(fit_obj, label) {
  d  <- get_fit_obs(fit_obj)
  y  <- d$obs; mu <- d$fit
  dr <- sign(y - mu) * sqrt(2 * pmax(
    y * log((y + 1e-8) / (mu + 1e-8)) - (y - mu), 0))
  data.frame(week = rep(d$wks, ncol(dr)),
             resid = as.vector(dr), model = label)
}

both <- !is.null(fit_m2a) && !is.null(fit_m2an)
```

## 9.1 Fitted vs Observed — City-Wide Weekly Total

```{r fitted-vs-obs, fig.height=4, fig.width=10}
if (both) {
  make_fit_df <- function(fit_obj, label) {
    d <- get_fit_obs(fit_obj)
    data.frame(week = d$wks, observed = rowSums(d$obs),
               fitted = rowSums(d$fit), model = label)
  }

  fit_df <- bind_rows(make_fit_df(fit_m2a,  "M2a"),
                      make_fit_df(fit_m2an, "M2a(N)"))

  ggplot(fit_df, aes(x = week)) +
    geom_line(aes(y = observed), color = "steelblue", linewidth = 0.7, alpha = 0.7) +
    geom_line(aes(y = fitted, color = model), linewidth = 0.8, linetype = "dashed") +
    geom_vline(xintercept = as.Date("2024-01-01"),
               color = "red", linetype = "dotted") +
    annotate("text", x = as.Date("2024-01-01"), y = Inf,
             label = "NFSU start", color = "red",
             vjust = 2, hjust = -0.1, size = 3) +
    scale_color_manual(values = c("M2a" = "darkgreen", "M2a(N)" = "darkorange")) +
    facet_wrap(~model) +
    labs(title = "Fitted vs Observed — City-Wide Weekly Total",
         subtitle = "Blue = observed | Dashed = fitted | Red dotted = NFSU start",
         x = "Week", y = "Total Shootings") +
    theme_minimal() + theme(legend.position = "none")
}
```

## 9.2 ACF / PACF of Pearson Residuals

```{r acf-pacf, fig.height=4}
plot_acf_pacf <- function(fit_obj, label) {
  if (is.null(fit_obj)) return(invisible(NULL))
  d        <- get_fit_obs(fit_obj)
  prs      <- (d$obs - d$fit) / sqrt(d$fit + 1e-8)
  resid_wk <- rowMeans(prs, na.rm = TRUE)
  op <- par(no.readonly = TRUE); on.exit(par(op), add = TRUE)
  par(mfrow = c(1, 2))
  acf(resid_wk,  main = paste("ACF —",  label), lag.max = 30, xlab = "Lag (weeks)")
  pacf(resid_wk, main = paste("PACF —", label), lag.max = 30, xlab = "Lag (weeks)")
}

plot_acf_pacf(fit_m2a,  "M2a")
plot_acf_pacf(fit_m2an, "M2a(N)")
```

## 9.3 Pearson Residual Distribution

```{r resid-dist, fig.height=4, fig.width=10}
if (both) {
  resid_df <- bind_rows(get_pearson_df(fit_m2a,  "M2a"),
                        get_pearson_df(fit_m2an, "M2a(N)"))

  ggplot(resid_df %>% filter(abs(resid) < 10), aes(x = resid, fill = model)) +
    geom_histogram(binwidth = 0.2, color = "white", alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
    facet_wrap(~model) +
    scale_fill_manual(values = c("M2a" = "darkgreen", "M2a(N)" = "darkorange")) +
    labs(title = "Pearson Residual Distribution — M2a vs M2a(N)",
         subtitle = "Clipped at |residual| < 10 for readability",
         x = "Pearson Residual", y = "Count") +
    theme_minimal() + theme(legend.position = "none")
}
```

## 9.4 QQ Plot of Pearson Residuals

```{r qq-plot, fig.height=4, fig.width=10}
if (both) {
  ggplot(resid_df %>% filter(abs(resid) < 10),
         aes(sample = resid, color = model)) +
    stat_qq(alpha = 0.3, size = 0.5) +
    stat_qq_line() +
    facet_wrap(~model) +
    scale_color_manual(values = c("M2a" = "darkgreen", "M2a(N)" = "darkorange")) +
    labs(title = "QQ Plot of Pearson Residuals — M2a vs M2a(N)",
         subtitle = "Closer to diagonal = better-specified model",
         x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal() + theme(legend.position = "none")
}
```

## 9.5 Mean Deviance Residuals Over Time

```{r deviance-resid, fig.height=4, fig.width=10}
if (both) {
  dev_df <- bind_rows(
    get_deviance_df(fit_m2a,  "M2a"),
    get_deviance_df(fit_m2an, "M2a(N)")
  ) %>%
    group_by(week, model) %>%
    summarize(mean_dev = mean(resid, na.rm = TRUE), .groups = "drop")

  ggplot(dev_df, aes(x = week, y = mean_dev, color = model)) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "grey50") +
    geom_hline(yintercept = 0, color = "grey30") +
    geom_vline(xintercept = as.Date("2024-01-01"),
               color = "red", linetype = "dotted") +
    scale_color_manual(values = c("M2a" = "darkgreen", "M2a(N)" = "darkorange")) +
    facet_wrap(~model) +
    labs(title = "Mean Deviance Residuals Over Time — M2a vs M2a(N)",
         subtitle = "Dashed bands at ±2 | Red dotted = NFSU start",
         x = "Week", y = "Mean deviance residual") +
    theme_minimal() + theme(legend.position = "none")
}
```

------------------------------------------------------------------------

# Section 10: Summary

```{r final-summary}
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║         M2a vs M2a(N) — Final Comparison Summary                ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")

cat("── Cluster Assignment ──────────────────────────────────────────────\n")
for (nm in names(cluster_list))
  cat(sprintf("  %-15s  %d grids\n", nm, length(cluster_list[[nm]])))
cat(sprintf("  %-15s  %d grids (reference — endemic intercept)\n\n",
    "Other", length(other_grids)))

cat("── Sparsity ────────────────────────────────────────────────────────\n")
cat(sprintf("  Overall panel: %.1f%% zeros\n\n",
    mean(panel$count == 0) * 100))

if (!is.null(fit_m2a) && !is.null(fit_m2an)) {
  cat("── Model Fit ───────────────────────────────────────────────────────\n")
  print(tbl)

  cat(sprintf("\n  LRT: χ² = %.4f  df = %d  p = %s\n\n",
      lrt, df_lrt, format(p_lrt, digits = 4)))

  c0 <- fit_m2a$coefficients
  cn <- fit_m2an$coefficients
  se0 <- sqrt(diag(fit_m2a$cov))
  sen <- sqrt(diag(fit_m2an$cov))

  # Recompute log-phi and SE for CIs
  lp0_b <- c0["ar.1"];  se0_b <- se0["ar.1"]
  lp0_a <- c0["ar.1"] + c0["ar.nfsu"]
  se0_a <- sqrt(se0["ar.1"]^2 + se0["ar.nfsu"]^2 + 2 * fit_m2a$cov["ar.1","ar.nfsu"])

  lpn_b <- cn["ar.1"];  sen_b <- sen["ar.1"]
  lpn_a <- cn["ar.1"] + cn["ar.nfsu"]
  sen_a <- sqrt(sen["ar.1"]^2 + sen["ar.nfsu"]^2 + 2 * fit_m2an$cov["ar.1","ar.nfsu"])

  cat("── φ Estimates with 95% CI (delta method) ──────────────────────────\n")
  cat(sprintf("  %-30s  φ_before = %.4f [%.4f, %.4f]\n",
      "M2a  (global)",
      exp(lp0_b), exp(lp0_b - 1.96*se0_b), exp(lp0_b + 1.96*se0_b)))
  cat(sprintf("  %-30s  φ_after  = %.4f [%.4f, %.4f]   Δ = %+.1f%%\n",
      "",
      exp(lp0_a), exp(lp0_a - 1.96*se0_a), exp(lp0_a + 1.96*se0_a),
      (exp(lp0_a) - exp(lp0_b)) / exp(lp0_b) * 100))
  cat(sprintf("  %-30s  φ_before = %.4f [%.4f, %.4f]\n",
      "M2a(N) (neighborhood)",
      exp(lpn_b), exp(lpn_b - 1.96*sen_b), exp(lpn_b + 1.96*sen_b)))
  cat(sprintf("  %-30s  φ_after  = %.4f [%.4f, %.4f]   Δ = %+.1f%%\n\n",
      "",
      exp(lpn_a), exp(lpn_a - 1.96*sen_a), exp(lpn_a + 1.96*sen_a),
      (exp(lpn_a) - exp(lpn_b)) / exp(lpn_b) * 100))

  cat("── Background Rate Offsets — M2a(N) ───────────────────────────────\n")
  cat(sprintf("  %-25s  Reference = Other (%d grids, endemic intercept)\n",
      "", length(other_grids)))
  for (nm in c("end.frogtown", "end.phalen", "end.mac_highland")) {
    lo_rr <- exp(cn[nm] - 1.96 * sen[nm])
    hi_rr <- exp(cn[nm] + 1.96 * sen[nm])
    cat(sprintf("  %-25s  log(RR)=%+.4f   RR=%.3f  [%.3f, %.3f]\n",
        nm, cn[nm], exp(cn[nm]), lo_rr, hi_rr))
  }

  cat(sprintf("\n  Convergence — M2a: %s | M2a(N): %s\n",
      fit_m2a$convergence, fit_m2an$convergence))
}
```
