---
title: "DASC 460: Analyzing the Spatial Spread of Violence in Saint Paul to Determine the Effectiveness of Establishing a Non-Fatal Shooting Unit"
output: html_document
---

------------------------------------------------------------------------

# Setup

## Load Libraries

```{r warning=FALSE, message=FALSE}
library(rmarkdown)
library(knitr)
library(igraph)
library(corrplot)
library(moments)
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(readr)
library(shiny)
library(geosphere)
library(tidyr)
```

## Load Data

```{r}
file <- "Shootings_Full Data(new).xlsx"
excel_sheets(file)

full_data <- read_excel(file, sheet = "Full Data")
loc_time  <- read_excel(file, sheet = "Loc_Time")
tables    <- read_excel(file, sheet = "Tables")
```

## Clean Data

```{r}
# Keep only rows with valid coordinates
map_data <- full_data %>%
  filter(!is.na(Lat), !is.na(Long), !is.na(DATE), Lat != 0, Long != 0) %>%
  mutate(
    DATE      = as.Date(DATE),
    Year      = format(DATE, "%Y"),
    month_day = format(DATE, "%m-%d")
  )
```

## Set Time Variables

```{r}
all_dates <- sort(unique(map_data$DATE))
years     <- sort(unique(map_data$Year))
years4    <- years[1:min(4, length(years))]
```

------------------------------------------------------------------------

# Section 1: Interactive Maps

## Visualization 1 - Point Map

Individual shooting incidents plotted as circle markers.

```{r}
leaflet(map_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~Long,
    lat         = ~Lat,
    radius      = 3,
    color       = "purple",
    opacity     = 0.8,
    fillOpacity = 0.4,
    popup = ~paste(
      "<b>CN:</b>", CN, "<br>",
      "<b>Date:</b>", as.Date(DATE), "<br>",
      "<b>Description:</b>", DESCRIPTION, "<br>",
      "<b>Disposition:</b>", DISPOSITION
    )
  )
```

## Visualization 2 - Cluster Aggregation Map

Markers grouped into clusters to show density by area.

```{r}
leaflet(map_data) %>%
  addTiles() %>%
  addMarkers(
    lng            = ~Long,
    lat            = ~Lat,
    clusterOptions = markerClusterOptions(showCoverageOnHover = TRUE)
  )
```

## Visualization 3 - Grid Aggregation Map

Circle size reflects the number of incidents within each grid cell.

```{r}
grid_counts <- full_data %>%
  filter(!is.na(GRID), !is.na(Lat), !is.na(Long)) %>%
  group_by(GRID) %>%
  summarize(
    count = n(),
    Lat   = mean(Lat),
    Long  = mean(Long)
  )

leaflet(grid_counts) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~Long,
    lat         = ~Lat,
    radius      = ~sqrt(count) * 2,
    color       = "red",
    fillOpacity = 0.5,
    popup = ~paste(
      "<b>GRID:</b>", GRID, "<br>",
      "<b>Count:</b>", count
    )
  )
```

## Visualization 4 - Animated Small Multiples (Shiny App)

Interactive Shiny app showing shooting progression before and after NFSU simultaneously. Use the slider or Play/Pause buttons to animate through time.

```{r}
# Use leap-year style full-year sequence so all month/day values exist
slider_days   <- seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day")
slider_labels <- format(slider_days, "%m-%d")

# Split data into Before/After NFSU periods
map_before <- map_data %>% filter(DATE < as.Date("2024-01-01"))
map_after  <- map_data %>% filter(DATE >= as.Date("2024-01-01"))

make_base_map <- function(df) {
  leaflet() %>%
    addTiles() %>%
    setView(
      lng  = mean(df$Long, na.rm = TRUE),
      lat  = mean(df$Lat,  na.rm = TRUE),
      zoom = 11
    )
}

ui <- fluidPage(
  titlePanel("Shootings Before vs After NFSU: Synchronized Time Progression"),

  fluidRow(
    column(3,
      tags$b("Progress through year:"),
      textOutput("date_label")
    ),
    column(3,
      actionButton("play",  "Play"),
      actionButton("pause", "Pause")
    ),
    column(6,
      sliderInput("speed", "Animation speed (smaller = faster)",
                  min = 100, max = 2000, value = 700, step = 100)
    )
  ),

  sliderInput(
    inputId = "day_index",
    label   = "Move through the calendar year",
    min     = 1,
    max     = length(slider_labels),
    value   = 1,
    step    = 1,
    animate = FALSE
  ),

  tabsetPanel(
    tabPanel("Points",
      fluidRow(
        column(6, h4("Before NFSU"), leafletOutput("points_before", height = 450)),
        column(6, h4("After NFSU"),  leafletOutput("points_after",  height = 450))
      )
    ),
    tabPanel("Clusters",
      fluidRow(
        column(6, h4("Before NFSU"), leafletOutput("clusters_before", height = 450)),
        column(6, h4("After NFSU"),  leafletOutput("clusters_after",  height = 450))
      )
    )
  )
)

server <- function(input, output, session) {

  playing <- reactiveVal(FALSE)

  observeEvent(input$play,  { playing(TRUE)  })
  observeEvent(input$pause, { playing(FALSE) })

  observe({
    if (playing()) {
      invalidateLater(input$speed, session)
      current <- input$day_index
      if (current < length(slider_labels)) {
        updateSliderInput(session, "day_index", value = current + 1)
      } else {
        playing(FALSE)
      }
    }
  })

  selected_month_day <- reactive({ slider_labels[input$day_index] })

  output$date_label <- renderText({ selected_month_day() })

  # Initialize blank base maps
  output$points_before   <- renderLeaflet({ make_base_map(map_data) })
  output$points_after    <- renderLeaflet({ make_base_map(map_data) })
  output$clusters_before <- renderLeaflet({ make_base_map(map_data) })
  output$clusters_after  <- renderLeaflet({ make_base_map(map_data) })

  # Helper: filter a period's data up to the selected month-day
  filtered_period_data <- function(df, month_day_cutoff) {
    df %>% filter(month_day <= month_day_cutoff)
  }

  # Points — Before NFSU
  observe({
    df <- filtered_period_data(map_before, selected_month_day())
    leafletProxy("points_before", data = df) %>%
      clearMarkers() %>%
      addCircleMarkers(
        lng         = ~Long,
        lat         = ~Lat,
        radius      = 3,
        color       = "red",
        opacity     = 0.8,
        fillOpacity = 0.4,
        popup = ~paste(
          "<b>Date:</b>", DATE, "<br>",
          "<b>CN:</b>", CN, "<br>",
          "<b>Description:</b>", DESCRIPTION, "<br>",
          "<b>Disposition:</b>", DISPOSITION
        )
      )
  })

  # Points — After NFSU
  observe({
    df <- filtered_period_data(map_after, selected_month_day())
    leafletProxy("points_after", data = df) %>%
      clearMarkers() %>%
      addCircleMarkers(
        lng         = ~Long,
        lat         = ~Lat,
        radius      = 3,
        color       = "blue",
        opacity     = 0.8,
        fillOpacity = 0.4,
        popup = ~paste(
          "<b>Date:</b>", DATE, "<br>",
          "<b>CN:</b>", CN, "<br>",
          "<b>Description:</b>", DESCRIPTION, "<br>",
          "<b>Disposition:</b>", DISPOSITION
        )
      )
  })

  # Clusters — Before NFSU
  observe({
    df <- filtered_period_data(map_before, selected_month_day())
    leafletProxy("clusters_before", data = df) %>%
      clearMarkers() %>%
      clearMarkerClusters() %>%
      addMarkers(
        lng   = ~Long,
        lat   = ~Lat,
        popup = ~paste(
          "<b>Date:</b>", DATE, "<br>",
          "<b>CN:</b>", CN, "<br>",
          "<b>Description:</b>", DESCRIPTION, "<br>",
          "<b>Disposition:</b>", DISPOSITION
        ),
        clusterOptions = markerClusterOptions(showCoverageOnHover = TRUE)
      )
  })

  # Clusters — After NFSU
  observe({
    df <- filtered_period_data(map_after, selected_month_day())
    leafletProxy("clusters_after", data = df) %>%
      clearMarkers() %>%
      clearMarkerClusters() %>%
      addMarkers(
        lng   = ~Long,
        lat   = ~Lat,
        popup = ~paste(
          "<b>Date:</b>", DATE, "<br>",
          "<b>CN:</b>", CN, "<br>",
          "<b>Description:</b>", DESCRIPTION, "<br>",
          "<b>Disposition:</b>", DISPOSITION
        ),
        clusterOptions = markerClusterOptions(showCoverageOnHover = TRUE)
      )
  })
}

shinyApp(ui, server)
```

------------------------------------------------------------------------

# Section 2: Temporal Patterns

## Prepare Temporal Variables

```{r}
map_data <- map_data %>%
  mutate(
    month  = floor_date(DATE, "month"),
    period = ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
    hour   = hour(Time),
    day    = wday(DATE, label = TRUE, week_start = 1)
  ) %>%
  mutate(
    period = factor(period, levels = c("Before NFSU", "After NFSU")),
    day    = factor(day,    levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
  )

monthly_counts <- map_data %>%
  group_by(month) %>%
  summarize(count = n(), .groups = "drop")
```

## Visualization 5.1 - Monthly Counts with NFSU Intervention Line

```{r}
ggplot(monthly_counts, aes(x = month, y = count)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  geom_vline(xintercept = as.Date("2024-01-01"), color = "red", linetype = "dashed") +
  annotate("text", x = as.Date("2024-01-01"), y = max(monthly_counts$count),
           label = "NFSU Start", color = "red", hjust = -0.1, size = 3) +
  labs(title = "Monthly Shootings Over Time", x = "Month", y = "Count")
```

## Visualization 5.2 - Yearly Counts (Bar)

```{r}
year_period_counts <- map_data %>%
  group_by(period, Year) %>%
  summarize(count = n(), .groups = "drop")

ggplot(year_period_counts, aes(x = Year, y = count)) +
  geom_col() +
  facet_wrap(~period, scales = "free_x") +
  labs(title = "Yearly Shootings: Before vs After NFSU", x = "Year", y = "Count")
```

## Visualization 5.3 - Before vs After NFSU (Bar)

```{r}
period_counts <- map_data %>%
  group_by(period) %>%
  summarize(count = n(), .groups = "drop")

ggplot(period_counts, aes(x = period, y = count, fill = period)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  labs(title = "Shootings Before vs After NFSU", x = "", y = "Count")
```

## Visualization 5.4 - Time of Day Pattern

```{r}
ggplot(map_data, aes(x = hour, fill = period)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  labs(title = "Shootings by Hour of Day: Before vs After NFSU",
       x = "Hour", y = "Count")
```

### Difference: Hour of Day (After - Before)

```{r}
hour_diff <- map_data %>%
  group_by(period, hour) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = count, values_fill = 0) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)

ggplot(hour_diff, aes(x = hour, y = diff, fill = diff > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title    = "Change in Shootings by Hour of Day (After - Before NFSU)",
    subtitle = "Red = increase, Blue = decrease",
    x        = "Hour of Day",
    y        = "Change in Count"
  ) +
  theme_minimal()
```

## Visualization 5.5 - Day of Week Pattern

```{r}
ggplot(map_data, aes(x = day, fill = period)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  labs(title = "Shootings by Day of Week: Before vs After NFSU", x = "Day", y = "Count")
```

### Difference: Day of Week (After - Before)

```{r}
day_diff <- map_data %>%
  group_by(period, day) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = count, values_fill = 0) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)

ggplot(day_diff, aes(x = day, y = diff, fill = diff > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title    = "Change in Shootings by Day of Week (After - Before NFSU)",
    subtitle = "Red = increase, Blue = decrease",
    x        = "Day of Week",
    y        = "Change in Count"
  ) +
  theme_minimal()
```

## Visualization 5.6 - Seasonal Patterns

```{r}
map_data <- map_data %>%
  mutate(
    season = case_when(
      month(DATE) %in% c(12, 1, 2)  ~ "Winter",
      month(DATE) %in% c(3, 4, 5)   ~ "Spring",
      month(DATE) %in% c(6, 7, 8)   ~ "Summer",
      month(DATE) %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  mutate(
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    period = factor(period, levels = c("Before NFSU", "After NFSU"))
  )

season_counts <- map_data %>%
  group_by(period, season) %>%
  summarize(count = n(), .groups = "drop")

ggplot(season_counts, aes(x = season, y = count, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  labs(title = "Seasonal Shootings: Before vs After NFSU", x = "Season", y = "Count") +
  theme_minimal()
```

### Difference: Season (After - Before)

```{r}
season_diff <- season_counts %>%
  pivot_wider(names_from = period, values_from = count, values_fill = 0) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)

ggplot(season_diff, aes(x = season, y = diff, fill = diff > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title    = "Change in Shootings by Season (After - Before NFSU)",
    subtitle = "Red = increase, Blue = decrease",
    x        = "Season",
    y        = "Change in Count"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 3: Spatial Patterns

## Visualization 5.7 - Spatial Density (Heat Map)

```{r}
ggplot(map_data, aes(x = Long, y = Lat)) +
  stat_density_2d(aes(fill = after_stat(level)), geom = "polygon") +
  scale_fill_viridis_c() +
  facet_wrap(~period) +
  labs(title = "Shooting Density Map: Before vs After NFSU", x = "Longitude", y = "Latitude") +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 4: Time Between Events

## Prepare Time-Gap Data

```{r}
shootings <- read_excel("Shootings_Full Data(new).xlsx", sheet = "Full Data")

shootings_gaps <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    )
  ) %>%
  filter(!is.na(event_dt)) %>%
  arrange(event_dt) %>%
  mutate(
    prev_event_dt = lag(event_dt),
    gap_minutes   = as.numeric(difftime(event_dt, prev_event_dt, units = "mins")),
    gap_hours     = as.numeric(difftime(event_dt, prev_event_dt, units = "hours")),
    gap_days      = as.numeric(difftime(event_dt, prev_event_dt, units = "days"))
  ) %>%
  filter(!is.na(gap_hours))
```

## Visualization 6.1 - Time Between Events: Hours

```{r}
ggplot(shootings_gaps %>% filter(gap_hours <= 168), aes(x = gap_hours)) +
  geom_histogram(binwidth = 4, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Time Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Hours Since Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Visualization 6.2 - Time Between Events: First 24 Hours

```{r}
ggplot(shootings_gaps %>% filter(gap_minutes <= 1440), aes(x = gap_minutes)) +
  geom_histogram(binwidth = 30, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Time Between Consecutive Shooting Events",
    subtitle = "First 24 Hours | Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Data Quality Check

```{r}
summary(shootings$Time)
sum(is.na(shootings$Time))
sum(is.na(shootings_gaps$event_dt))
table(shootings_gaps$period)
nrow(shootings_gaps)
```

------------------------------------------------------------------------

# Section 5: Distance Between Events

## Prepare Distance Data

```{r}
shootings_space <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    )
  ) %>%
  filter(!is.na(event_dt), !is.na(Lat), !is.na(Long)) %>%
  arrange(event_dt) %>%
  mutate(
    prev_lat    = lag(Lat),
    prev_long   = lag(Long),
    dist_meters = distHaversine(
      matrix(c(prev_long, prev_lat), ncol = 2),
      matrix(c(Long, Lat),           ncol = 2)
    ),
    dist_miles = dist_meters / 1609.34,
    dist_feet  = dist_meters * 3.28084
  ) %>%
  filter(!is.na(dist_miles))
```

## Visualization 7.1 - Distance Between Events: Miles

```{r}
ggplot(shootings_space %>% filter(dist_miles <= 5), aes(x = dist_miles)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distance Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Miles from Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Visualization 7.2 - Distance Between Events: Miles (Normalized)

Normalized to compare the *shape* of the spatial spread pattern independent of volume - useful for detecting whether shooting geography changed after NFSU.

```{r}
ggplot(shootings_space %>% filter(dist_miles <= 5),
       aes(x = dist_miles, y = after_stat(density))) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distribution of Distance Between Events",
    subtitle = "Normalized Comparison | Before vs After NFSU",
    x        = "Miles from Previous Event",
    y        = "Density"
  ) +
  theme_minimal()
```

## Visualization 7.3 - Distance Between Events: Feet

```{r}
ggplot(shootings_space %>% filter(dist_feet <= 10000), aes(x = dist_feet)) +
  geom_histogram(binwidth = 500, fill = "steelblue", color = "white") +
  facet_wrap(~period, scales = "fixed") +
  labs(
    title    = "Distance Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Feet from Previous Event",
    y        = "Count"
  ) +
  theme_minimal()
```

## Summary Statistics - Distance

```{r}
shootings_space %>%
  group_by(period) %>%
  summarize(
    n                 = n(),
    mean_dist_miles   = mean(dist_miles,   na.rm = TRUE),
    median_dist_miles = median(dist_miles, na.rm = TRUE),
    sd_dist_miles     = sd(dist_miles,     na.rm = TRUE),
    min_dist_miles    = min(dist_miles,    na.rm = TRUE),
    max_dist_miles    = max(dist_miles,    na.rm = TRUE)
  )
```

------------------------------------------------------------------------

# Section 6: Space-Time Heatmaps

## Prepare Day × Hour Heatmap Data

```{r}
shootings_time <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = if_else(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
    hour     = hour(event_dt),
    day      = wday(event_dt, label = TRUE, week_start = 1)
  ) %>%
  filter(!is.na(event_dt)) %>%
  mutate(
    day    = factor(day,    levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
    period = factor(period, levels = c("Before NFSU", "After NFSU"))
  )

heatmap_data <- shootings_time %>%
  group_by(period, day, hour) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(
    period = factor(period, levels = c("Before NFSU", "After NFSU")),
    day    = factor(day,    levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
  )
```

## Visualization 8.1 - Day × Hour Heatmap (Raw Counts)

```{r}
ggplot(heatmap_data, aes(x = hour, y = day, fill = count)) +
  geom_tile() +
  facet_wrap(~period) +
  scale_fill_viridis_c() +
  labs(
    title    = "Shootings by Day of Week and Hour",
    subtitle = "Before vs After NFSU",
    x        = "Hour of Day",
    y        = "Day of Week",
    fill     = "Count"
  ) +
  theme_minimal()
```

## Visualization 8.2 - Day × Hour Heatmap (Normalized)

Normalized to compare *when* shootings cluster across the week — useful for spotting pattern shifts independent of volume differences.

```{r}
heatmap_data_norm <- heatmap_data %>%
  group_by(period) %>%
  mutate(prop = count / sum(count))

ggplot(heatmap_data_norm, aes(x = hour, y = day, fill = prop)) +
  geom_tile() +
  facet_wrap(~period) +
  scale_fill_viridis_c() +
  labs(
    title    = "Distribution of Shootings by Day & Hour",
    subtitle = "Normalized | Before vs After NFSU",
    x        = "Hour of Day",
    y        = "Day of Week",
    fill     = "Proportion"
  ) +
  theme_minimal()
```

## Visualization 8.3 - Day × Hour Difference Heatmap (After - Before)

```{r}
diff_data <- heatmap_data %>%
  pivot_wider(names_from = period, values_from = count, values_fill = 0) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)

ggplot(diff_data, aes(x = hour, y = day, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title    = "Change in Shootings by Day & Hour (After - Before NFSU)",
    subtitle = "Red = increase, Blue = decrease",
    x        = "Hour",
    y        = "Day",
    fill     = "Change"
  ) +
  theme_minimal()
```

## Visualization 8.4 - Hourly Patterns by Day (Facet Grid)

```{r}
heatmap_data %>%
  ggplot(aes(x = hour, y = count)) +
  geom_line(color = "steelblue") +
  facet_grid(day ~ period) +
  labs(
    title = "Hourly Shooting Patterns by Day",
    x     = "Hour of Day",
    y     = "Count"
  )
```

------------------------------------------------------------------------

# Section 7: Joint Space-Time Analysis

## Prepare Space-Time Data

```{r}
space_time <- shootings %>%
  mutate(
    DATE     = as.Date(DATE),
    Time     = format(as.POSIXct(Time, format = "%H:%M:%S"), "%H:%M:%S"),
    event_dt = ymd_hms(paste(DATE, Time)),
    period   = if_else(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU")
  ) %>%
  filter(!is.na(event_dt), !is.na(Lat), !is.na(Long)) %>%
  arrange(event_dt) %>%
  mutate(
    prev_event_dt = lag(event_dt),
    prev_lat      = lag(Lat),
    prev_long     = lag(Long),
    gap_minutes   = as.numeric(difftime(event_dt, prev_event_dt, units = "mins")),
    dist_meters   = distHaversine(
      matrix(c(prev_long, prev_lat), ncol = 2),
      matrix(c(Long, Lat),           ncol = 2)
    ),
    dist_miles = dist_meters / 1609.34
  ) %>%
  filter(!is.na(gap_minutes), !is.na(dist_miles)) %>%
  mutate(period = factor(period, levels = c("Before NFSU", "After NFSU")))
```

## Visualization 9.1 - Time vs Distance (2D Bin)

```{r}
ggplot(space_time %>% filter(gap_minutes <= 1440, dist_miles <= 5),
       aes(x = gap_minutes, y = dist_miles)) +
  geom_bin2d() +
  facet_wrap(~period) +
  labs(
    title    = "Time and Distance Between Consecutive Shooting Events",
    subtitle = "First 24 Hours and Within 5 Miles | Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Miles From Previous Event",
    fill     = "Count"
  ) +
  theme_minimal()
```

### Difference: 2D Bin (After - Before)

```{r}
bin_counts <- space_time %>%
  filter(gap_minutes <= 1440, dist_miles <= 5) %>%
  mutate(
    gap_bin  = cut(gap_minutes, breaks = seq(0, 1440, by = 60),  include.lowest = TRUE),
    dist_bin = cut(dist_miles,  breaks = seq(0, 5,    by = 0.5), include.lowest = TRUE)
  ) %>%
  group_by(period, gap_bin, dist_bin) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = count, values_fill = 0) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)

ggplot(bin_counts, aes(x = gap_bin, y = dist_bin, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title    = "Change in Event Pairs by Time & Distance (After - Before NFSU)",
    subtitle = "Red = more pairs After NFSU, Blue = fewer pairs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Miles From Previous Event",
    fill     = "Change"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Visualization 9.2 - Time vs Distance (Scatter)

```{r}
ggplot(space_time %>% filter(gap_minutes <= 1440, dist_miles <= 5),
       aes(x = gap_minutes, y = dist_miles)) +
  geom_point(alpha = 0.25, size = 1) +
  facet_wrap(~period) +
  labs(
    title    = "Time and Distance Between Consecutive Shooting Events",
    subtitle = "Before vs After NFSU",
    x        = "Minutes Since Previous Event",
    y        = "Miles From Previous Event"
  ) +
  theme_minimal()
```

# Section 9: Summary Statistics Table (Before vs After NFSU)

## Prepare disposition data

```{r}
disposition_summary <- full_data %>%
  mutate(
    DATE   = as.Date(DATE),
    period = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    )
  ) %>%
  filter(!is.na(DISPOSITION)) %>%
  group_by(period, DISPOSITION) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(period) %>%
  mutate(pct = round(100 * count / sum(count), 1)) %>%
  arrange(period, desc(count))
```

## Visualization 10.1 - Disposition breakdown Before vs After NFSU

```{r}
ggplot(disposition_summary, aes(x = reorder(DISPOSITION, count), y = pct, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  coord_flip() +
  labs(
    title    = "Case Disposition Breakdown: Before vs After NFSU",
    subtitle = "As % of total cases in each period",
    x        = "Disposition",
    y        = "Percentage (%)",
    fill     = "Period"
  ) +
  theme_minimal()
```

## Visualization 10.2 - Overall summary stats table

```{r}
summary_table <- map_data %>%
  group_by(period) %>%
  summarize(
    total_incidents      = n(),
    date_range           = paste(min(DATE), "to", max(DATE)),
    avg_per_month        = round(n() / n_distinct(format(DATE, "%Y-%m")), 1),
    pct_weekend          = round(100 * mean(wday(DATE) %in% c(1, 7)), 1),
    pct_summer           = round(100 * mean(month(DATE) %in% c(6, 7, 8)), 1),
    pct_late_night       = round(100 * mean(hour(Time) %in% c(21, 22, 23, 0, 1, 2), na.rm = TRUE), 1),
    .groups              = "drop"
  )

kable(
  summary_table,
  col.names = c("Period", "Total Incidents", "Date Range",
                "Avg/Month", "% Weekend", "% Summer", "% Late Night (9pm–2am)"),
  caption   = "Summary Statistics: Before vs After NFSU"
)
```

## Visualization 10.3 - Top grid cells Before vs After (potential covariates)

```{r}
top_grids <- full_data %>%
  mutate(
    DATE   = as.Date(DATE),
    period = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    )
  ) %>%
  filter(!is.na(GRID)) %>%
  group_by(period, GRID) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(period) %>%
  slice_max(count, n = 10) %>%
  ungroup() %>%
  mutate(GRID = factor(GRID))
```

```{r}
ggplot(top_grids, aes(x = reorder(GRID, count), y = count, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  coord_flip() +
  labs(
    title    = "Top 10 Grid Cells by Incident Count: Before vs After NFSU",
    subtitle = "Identifies persistent spatial hotspots — candidates for background intensity covariates",
    x        = "Grid Cell",
    y        = "Count",
    fill     = "Period"
  ) +
  theme_minimal()
```

# Section 10: Case Resolution Time (DATE CLOSED - DATE)

```{r}
resolution <- full_data %>%
  mutate(
    DATE        = as.Date(DATE),
    DATE_CLOSED = as.Date(`DATE CLOSED`),
    period      = factor(
      ifelse(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
      levels = c("Before NFSU", "After NFSU")
    ),
    days_to_close = as.numeric(DATE_CLOSED - DATE)
  ) %>%
  filter(!is.na(days_to_close), days_to_close >= 0)
```

## Visualization 10.1 - Distribution of Resolution Time

```{r}
ggplot(resolution %>% filter(days_to_close <= 365),
       aes(x = days_to_close, fill = period)) +
  geom_histogram(binwidth = 14, position = "identity", alpha = 0.6, color = "white") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  facet_wrap(~period) +
  labs(
    title    = "Case Resolution Time: Before vs After NFSU",
    subtitle = "Days from incident date to date closed | Within 1 year",
    x        = "Days to Close",
    y        = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Visualization 10.2 - Median resolution time by disposition

```{r}
resolution_by_disp <- resolution %>%
  filter(days_to_close <= 365) %>%
  group_by(period, DISPOSITION) %>%
  summarize(
    median_days = median(days_to_close, na.rm = TRUE),
    mean_days   = mean(days_to_close,   na.rm = TRUE),
    n           = n(),
    .groups     = "drop"
  ) %>%
  filter(n >= 5)  # drop very rare disposition categories

ggplot(resolution_by_disp,
       aes(x = reorder(DISPOSITION, median_days), y = median_days, fill = period)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Before NFSU" = "red", "After NFSU" = "blue")) +
  coord_flip() +
  labs(
    title    = "Median Days to Close by Disposition: Before vs After NFSU",
    subtitle = "Only dispositions with 5+ cases shown",
    x        = "Disposition",
    y        = "Median Days to Close",
    fill     = "Period"
  ) +
  theme_minimal()
```

## Visualization 10.3 - Difference plot (After - Before median days)

```{r}
resolution_diff <- resolution_by_disp %>%
  select(period, DISPOSITION, median_days) %>%
  pivot_wider(names_from = period, values_from = median_days, values_fill = NA) %>%
  filter(!is.na(`Before NFSU`), !is.na(`After NFSU`)) %>%
  mutate(diff = `After NFSU` - `Before NFSU`)
ggplot(resolution_diff, aes(x = reorder(DISPOSITION, diff), y = diff, fill = diff > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue")) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(
    title    = "Change in Median Days to Close by Disposition (After - Before NFSU)",
    subtitle = "Red = took longer after NFSU | Blue = resolved faster after NFSU",
    x        = "Disposition",
    y        = "Change in Median Days"
  ) +
  theme_minimal()
```

## Summary stats table

```{r}
resolution_summary <- resolution %>%
  filter(days_to_close <= 365) %>%
  group_by(period) %>%
  summarize(
    n              = n(),
    mean_days      = round(mean(days_to_close,   na.rm = TRUE), 1),
    median_days    = round(median(days_to_close, na.rm = TRUE), 1),
    pct_closed_7d  = round(100 * mean(days_to_close <= 7,   na.rm = TRUE), 1),
    pct_closed_30d = round(100 * mean(days_to_close <= 30,  na.rm = TRUE), 1),
    pct_closed_90d = round(100 * mean(days_to_close <= 90,  na.rm = TRUE), 1),
    .groups        = "drop"
  )

kable(
  resolution_summary,
  col.names = c("Period", "n", "Mean Days", "Median Days",
                "% Closed ≤7 Days", "% Closed ≤30 Days", "% Closed ≤90 Days"),
  caption   = "Case Resolution Time Summary: Before vs After NFSU"
)
```
