---
title: "St. Paul Shootings: Background Intensity Model"
subtitle: "Poisson GLM for the Hawkes Process Background Rate μ(t)"
author: "DASC 460 / STAT 460"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

# Purpose and Context

This file models the **background intensity function μ(t)** of a Hawkes
(self-exciting) point process applied to non-fatal shooting incidents in
St. Paul, MN (2022–2025).

In a Hawkes process, the total conditional intensity (expected rate of
events at time t) is:

$$\lambda(t) = \underbrace{\mu(t)}_{\text{background rate}} + \underbrace{\sum_{t_i < t} g(t - t_i)}_{\text{self-excitation from past events}}$$

**This file focuses entirely on estimating μ(t)** — the baseline
shooting rate driven by measurable time-varying factors — before any
self-excitation is modeled.

We estimate μ(t) using a **Poisson generalized linear model (GLM)** with
a log link, where the response is the daily count of shooting incidents
and the predictors are time-based covariates identified from exploratory
analysis and Dr. Clark's guidance.

The key substantive question throughout: **Did the Non-Fatal Shooting
Unit (NFSU), established January 1, 2024, produce a measurable reduction
in shootings beyond the pre-existing trend?**

------------------------------------------------------------------------

# Setup: Libraries and Data

```{r libraries}
library(readxl)    # reading Excel data
library(dplyr)     # data manipulation
library(lubridate) # date/time handling
library(ggplot2)   # visualization
library(MASS)      # glm.nb() for negative binomial
```

```{r load-data}
# Load the shooting incident data (one row per incident)
shootings_raw <- read_excel("Shootings_Full Data(new).xlsx", sheet = "Full Data") %>%
  mutate(DATE = as.Date(DATE))

cat("Total shooting incidents in dataset:", nrow(shootings_raw), "\n")
cat("Date range:", as.character(min(shootings_raw$DATE)),
    "to", as.character(max(shootings_raw$DATE)), "\n")
```

------------------------------------------------------------------------

# Building the Daily Count Panel

## Why a complete daily panel?

The raw data only contains rows for days when shootings occurred. For
Poisson regression, we need **every day in the study window** —
including days with zero incidents — so the model correctly accounts for
the exposure time when incidents did *not* occur.

Without zero-count days, the model would be estimated only on "active"
days and would systematically overestimate the background rate.

```{r build-panel}
# Step 1: Create a row for every calendar day in the study window
all_study_days <- data.frame(
  DATE = seq(min(shootings_raw$DATE), max(shootings_raw$DATE), by = "day")
)

# Step 2: Count incidents per day from raw data
daily_incident_counts <- shootings_raw %>%
  group_by(DATE) %>%
  summarise(shooting_count = n(), .groups = "drop")

# Step 3: Join counts onto the full day sequence; zero-fill missing days
daily_data <- all_study_days %>%
  left_join(daily_incident_counts, by = "DATE") %>%
  mutate(shooting_count = replace(shooting_count, is.na(shooting_count), 0)) %>%
  arrange(DATE)

cat("Complete daily panel:\n")
cat("  Total days in panel:", nrow(daily_data), "\n")
cat("  Days with zero incidents:", sum(daily_data$shooting_count == 0), "\n")
cat("  Days with 1+ incidents:", sum(daily_data$shooting_count > 0), "\n")
cat("  Max incidents in a single day:", max(daily_data$shooting_count), "\n")
```

------------------------------------------------------------------------

# Covariate Engineering

Each covariate below was motivated by either exploratory analysis
(visualizations showed clear patterns) or Dr. Clark's model
specification from meetings.

```{r covariates}
# Small constant to allow safe log-transformation of zero-count lag values
LOG_ZERO_EPSILON <- 1e-6

daily_data <- daily_data %>%
  mutate(

    # ── TREND ──────────────────────────────────────────────────────────────────
    # How it was found: Monthly count plots showed a clear downward trend in
    # shootings from 2022 through 2025. Dr. Clark explicitly specified a linear
    # time trend as a required covariate (Stat meet 3).
    # Construction: Integer index starting at 0 on the first day of the study.
    # In the Poisson model: captures the average day-over-day change in the
    # log-expected shooting rate, independent of all other factors.
    time_trend = as.numeric(DATE - min(DATE)),

    # ── NFSU INDICATOR ─────────────────────────────────────────────────────────
    # How it was found: The Non-Fatal Shooting Unit became operational on
    # January 1, 2024 (confirmed in the police department visit). This is the
    # primary intervention variable of interest for the project.
    # Construction: Binary 0/1 — 0 before NFSU, 1 on or after Jan 1, 2024.
    # In the model: estimates the level shift in log-expected rate associated
    # with the unit's establishment, after controlling for trend and timing.
    nfsu_active = ifelse(DATE >= as.Date("2024-01-01"), 1, 0),

    # ── WEEKEND INDICATOR ──────────────────────────────────────────────────────
    # How it was found: Day-of-week bar charts and day×hour heatmaps showed
    # consistently higher shooting counts on Saturdays and Sundays across all
    # years. Dr. Clark included this in the baseline model specification.
    # Construction: Binary 1 = Saturday or Sunday, 0 = weekday.
    # In the model: captures the systematic weekly rhythmicity in shooting risk.
    is_weekend = ifelse(wday(DATE) %in% c(1, 7), 1, 0),

    # ── SEASON ─────────────────────────────────────────────────────────────────
    # How it was found: Monthly and seasonal bar charts showed a pronounced
    # winter dip in shootings, consistent with the literature on seasonal
    # crime patterns. Spring and summer were elevated but less distinctly so.
    # Construction: Four meteorological seasons based on calendar month.
    # Fall is the reference category in the model.
    # In the model: captures systematic seasonal variation in the background rate.
    season = case_when(
      month(DATE) %in% c(12, 1, 2) ~ "Winter",
      month(DATE) %in% c(3, 4, 5)  ~ "Spring",
      month(DATE) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                          ~ "Fall"          # reference level
    ),

    # ── INTERACTION: WEEKEND × NFSU ────────────────────────────────────────────
    # How it was found: Dr. Clark's baseline model specification included this
    # term to test whether the NFSU changed the *weekend effect*, not just the
    # overall level. Theoretically, if the unit altered patrol patterns on
    # weekends specifically, β_weekend:nfsu would be non-zero.
    # (Computed implicitly by including weekend:nfsu in the glm() formula)

    # ── LAGGED COUNT (self-excitation proxy) ───────────────────────────────────
    # How it was found: Dr. Clark introduced this at DASC meet 4 as a discrete
    # approximation to the Hawkes self-excitation term. The idea: if yesterday
    # had many shootings, today's expected count may be elevated (repeat
    # victimization / retaliation dynamics). The lag is logged so it enters
    # the log-linear model on the right scale. A small epsilon is added before
    # logging to handle zero-count days without producing -Inf.
    # Construction: Previous day's shooting count, log-transformed.
    lag_shooting_count     = lag(shooting_count, n = 1, default = 0),
    log_lag_shooting_count = log(lag_shooting_count + LOG_ZERO_EPSILON)
  )

cat("Covariate summary:\n")
cat("  Trend range: 0 to", max(daily_data$time_trend), "(days)\n")
cat("  NFSU active days:", sum(daily_data$nfsu_active), "\n")
cat("  Weekend days:", sum(daily_data$is_weekend), "\n")
cat("  Season distribution:\n")
print(table(daily_data$season))
```

------------------------------------------------------------------------

# Model Fitting

We fit four nested Poisson GLMs to build from Clark's baseline to our
best model, plus a robustness check using a negative binomial
distribution.

All models use a **log link**, so coefficients are interpreted as
additive effects on log(μ) — equivalently, multiplicative effects on μ
itself after exponentiation.

The model family:
$$Y_t \sim \text{Poisson}(\mu_t), \quad \log(\mu_t) = \mathbf{X}_t \boldsymbol{\beta}$$

```{r fit-models}
# ── M0: Clark's exact baseline specification (Stat meet 3 whiteboard) ──────────
# log(μ_t) = β₀ + β₁·trend + β₂·weekend + β₃·nfsu + β₄·(weekend × nfsu)
m0_baseline <- glm(
  shooting_count ~ time_trend + is_weekend + nfsu_active + is_weekend:nfsu_active,
  family = poisson(link = "log"),
  data   = daily_data
)

# ── M1: Add season (motivated by exploratory visualizations) ──────────────────
# log(μ_t) = β₀ + β₁·trend + β₂·weekend + β₃·nfsu + β₄·(weekend×nfsu) + β_season
m1_with_season <- glm(
  shooting_count ~ time_trend + is_weekend + nfsu_active + is_weekend:nfsu_active
                 + factor(season),
  family = poisson(link = "log"),
  data   = daily_data
)

# ── M2: Add logged lag (discrete self-excitation proxy, DASC meet 4) ──────────
# log(μ_t) = [M1 terms] + β·log(y_{t-1} + ε)
m2_with_lag <- glm(
  shooting_count ~ time_trend + is_weekend + nfsu_active + is_weekend:nfsu_active
                 + factor(season) + log_lag_shooting_count,
  family = poisson(link = "log"),
  data   = daily_data
)

# ── M3: Drop trend — LRT test requested by Dr. Clark (DASC meet 4) ────────────
# Tests whether trend is necessary once season and lag are included
m3_no_trend <- glm(
  shooting_count ~ is_weekend + nfsu_active + is_weekend:nfsu_active
                 + factor(season) + log_lag_shooting_count,
  family = poisson(link = "log"),
  data   = daily_data
)

# ── M1-NB: Negative Binomial robustness check for mild overdispersion ─────────
# Dispersion ~1.28 in M1 suggests mild overdispersion; NB adds a free
# dispersion parameter θ to account for this
m1_negbin <- glm.nb(
  shooting_count ~ time_trend + is_weekend + nfsu_active + is_weekend:nfsu_active
                 + factor(season),
  data = daily_data
)

cat("All models fitted successfully.\n")
```

------------------------------------------------------------------------

# Model Comparison

## AIC / BIC

Lower AIC and BIC indicate better model fit, penalized for model
complexity.

```{r model-comparison-aic}
aic_table <- AIC(m0_baseline, m1_with_season, m2_with_lag, m3_no_trend)
bic_table <- BIC(m0_baseline, m1_with_season, m2_with_lag, m3_no_trend)

comparison_table <- data.frame(
  Model       = c("M0: Baseline (trend+weekend+nfsu)",
                  "M1: + Season",
                  "M2: + Season + Log Lag",
                  "M3: No Trend (+ Season + Log Lag)"),
  df          = aic_table$df,
  AIC         = round(aic_table$AIC, 1),
  BIC         = round(bic_table$BIC, 1)
)

knitr::kable(comparison_table, caption = "Model Comparison: AIC and BIC")
```

## Likelihood Ratio Tests

LRT tests whether each added term significantly improves fit.

```{r lrt-tests}
cat("=== LRT: M0 → M1  (Does adding season help?) ===\n")
print(anova(m0_baseline, m1_with_season, test = "Chisq"))

cat("\n=== LRT: M1 → M2  (Does the lag term add anything?) ===\n")
print(anova(m1_with_season, m2_with_lag, test = "Chisq"))

cat("\n=== LRT: M3 → M2  (Is trend necessary?) ===\n")
print(anova(m3_no_trend, m2_with_lag, test = "Chisq"))
```

**Interpretation:**

-   Season significantly improves fit over M0 (p \< 0.001) — include it.
-   The logged lag does **not** significantly improve fit over M1 (p ≈
    0.27) — at daily granularity, yesterday's count adds little once
    trend and season are controlled for.
-   Trend is highly significant (p \< 2e-16) — cannot be dropped.

**→ M1 is the best model by AIC, BIC, and LRT.**

## Dispersion Check

```{r dispersion}
pearson_dispersion <- function(model) {
  sum(residuals(model, type = "pearson")^2) / model$df.residual
}

disp_table <- data.frame(
  Model      = c("M0", "M1", "M2", "M1-NB"),
  Dispersion = round(c(pearson_dispersion(m0_baseline),
                       pearson_dispersion(m1_with_season),
                       pearson_dispersion(m2_with_lag),
                       pearson_dispersion(m1_negbin)), 3),
  Note = c("Mild overdispersion",
            "Mild overdispersion",
            "Mild overdispersion",
            "Free dispersion parameter θ")
)
knitr::kable(disp_table,
             caption = "Pearson Dispersion Statistic (ideal ≈ 1.0)")
```

Dispersion \~1.28 in the Poisson models indicates mild overdispersion.
The Negative Binomial (M1-NB) addresses this and provides a useful
robustness check.

------------------------------------------------------------------------

# Results: Best Model (M1)

## Coefficient Table

```{r m1-summary}
summary(m1_with_season)
```

## Exponentiated Coefficients (Incidence Rate Ratios)

Exponentiating the log-scale coefficients gives **Incidence Rate Ratios
(IRRs)**: values \> 1 increase expected shootings, values \< 1 decrease
them, relative to baseline.

```{r irr-table}
irr_estimates <- exp(coef(m1_with_season))
irr_ci        <- exp(confint(m1_with_season))

irr_table <- round(cbind(IRR = irr_estimates, irr_ci), 3)
rownames(irr_table) <- c("Intercept (Fall, weekday, pre-NFSU, trend=0)",
                          "Time Trend (per day)",
                          "Weekend vs Weekday",
                          "NFSU Active vs Inactive",
                          "Spring vs Fall",
                          "Summer vs Fall",
                          "Winter vs Fall",
                          "Weekend × NFSU Interaction")

knitr::kable(irr_table,
             col.names = c("IRR", "95% CI Lower", "95% CI Upper"),
             caption   = "M1 Incidence Rate Ratios (exponentiated coefficients)")
```

## Negative Binomial Robustness Check

Coefficients should be essentially identical to M1-Poisson if the
Poisson specification is appropriate. Differences in significance
indicate sensitivity to the overdispersion assumption.

```{r nb-summary}
cat("Poisson M1 AIC:", round(AIC(m1_with_season), 1), "\n")
cat("NegBin  M1 AIC:", round(AIC(m1_negbin), 1), "\n\n")
summary(m1_negbin)
```

------------------------------------------------------------------------

# Diagnostics

## Deviance Residuals Over Time

Residuals should appear as random noise around zero with no systematic
patterns. Persistent structure (trends, seasonal bulges, etc.) would
indicate missing covariates.

```{r residuals-over-time, fig.width=11, fig.height=4}
plot_residuals_over_time <- function(model, model_label) {
  daily_data %>%
    mutate(deviance_residual = residuals(model, type = "deviance")) %>%
    ggplot(aes(x = DATE, y = deviance_residual)) +
    geom_line(alpha = 0.45, linewidth = 0.35) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 0.7) +
    geom_vline(xintercept = as.Date("2024-01-01"),
               color = "steelblue", linetype = "dotted", linewidth = 0.7) +
    annotate("text", x = as.Date("2024-02-01"), y = 4.2,
             label = "NFSU Start", color = "steelblue", hjust = 0, size = 3) +
    labs(title = paste(model_label, "— Deviance Residuals Over Time"),
         subtitle = "No systematic pattern = model captures main temporal structure",
         x = "Date", y = "Deviance Residual") +
    theme_minimal()
}

plot_residuals_over_time(m0_baseline,    "M0 (Baseline)")
plot_residuals_over_time(m1_with_season, "M1 (Best Model)")
```

## ACF and PACF of Deviance Residuals

If the model has captured all temporal structure, residuals should be
uncorrelated across time lags — ACF and PACF bars should fall within the
blue confidence bands. Significant spikes would indicate remaining
autocorrelation that the model misses.

```{r acf-pacf, fig.width=11, fig.height=4}
plot_acf_pacf <- function(model, model_label) {
  res <- residuals(model, type = "deviance")
  par(mfrow = c(1, 2))
  acf(res,  main = paste(model_label, "— ACF"),  lag.max = 30)
  pacf(res, main = paste(model_label, "— PACF"), lag.max = 30)
  par(mfrow = c(1, 1))
}

plot_acf_pacf(m0_baseline,    "M0 (Baseline)")
plot_acf_pacf(m1_with_season, "M1 (Best Model)")
```

**ACF/PACF interpretation:** Lags are mostly within bounds for both M0
and M1. The PACF shows a few small spikes (max \~0.08) but these are
near the significance threshold and lack a systematic pattern. The model
captures the main temporal structure adequately for a background
intensity estimator.

------------------------------------------------------------------------

# Fitted Values vs Observed Counts

```{r fitted-vs-observed, fig.width=12, fig.height=5}
daily_data %>%
  mutate(
    fitted_m0 = fitted(m0_baseline),
    fitted_m1 = fitted(m1_with_season)
  ) %>%
  ggplot(aes(x = DATE)) +
  geom_point(aes(y = shooting_count), alpha = 0.2, size = 0.8, color = "grey40") +
  geom_line(aes(y = fitted_m1, color = "M1 (Best)"), linewidth = 0.8) +
  geom_vline(xintercept = as.Date("2024-01-01"),
             linetype = "dashed", color = "steelblue", linewidth = 0.7) +
  annotate("text", x = as.Date("2024-02-01"),
           y = max(daily_data$shooting_count) * 0.88,
           label = "NFSU Start (Jan 1, 2024)",
           color = "steelblue", hjust = 0, size = 3) +
  scale_color_manual(values = c("M1 (Best)" = "red")) +
  labs(title = "M1 Best Model: Observed Daily Shooting Counts vs Fitted Background Rate",
       subtitle = paste("The red line is μ̂(t) — the estimated Poisson background rate.",
                        "The sawtooth pattern reflects the weekend effect riding on a downward trend."),
       x = "Date", y = "Daily Shooting Count",
       color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The fitted line (μ̂) oscillates between weekday and weekend predicted
rates on top of the long-run downward trend. Extreme event days (counts
well above the line) are the "spikes" that a Hawkes self-excitation term
would be tasked with explaining.

------------------------------------------------------------------------

# Summary of Findings

```{r findings-table}
findings <- data.frame(
  Covariate = c(
    "Time Trend",
    "Weekend (Sat/Sun vs weekday)",
    "NFSU Active (vs pre-NFSU)",
    "Winter (vs Fall)",
    "Spring (vs Fall)",
    "Summer (vs Fall)",
    "Weekend × NFSU Interaction",
    "Logged Prior-Day Count (lag)"
  ),
  IRR = c("0.999", "1.31", "1.01", "0.73", "0.97", "1.02", "0.93", "—"),
  `95% CI` = c("(0.999, 0.999)", "(1.20, 1.44)", "(0.85, 1.19)",
                "(0.65, 0.82)",  "(0.87, 1.08)", "(0.92, 1.14)",
                "(0.78, 1.10)",  "—"),
  Significant = c("Yes (p < 2e-16)", "Yes (p < 0.001)", "No (p = 0.947)",
                   "Yes (p < 0.001)", "No (p = 0.607)", "No (p = 0.678)",
                   "No (p = 0.379)", "No (p = 0.27, LRT)"),
  Interpretation = c(
    "Shootings declining ~0.13%/day over study period",
    "~31% more shootings on weekends",
    "No significant level change after NFSU launch",
    "~27% fewer shootings in winter",
    "Not distinguishable from fall",
    "Not distinguishable from fall",
    "Weekend pattern unchanged post-NFSU",
    "Adds nothing once trend + season included"
  )
)

knitr::kable(findings,
             col.names = c("Covariate", "IRR", "95% CI",
                            "Significant?", "Plain-English Interpretation"),
             caption = "M1 (Best Model) — Complete Results Summary")
```

------------------------------------------------------------------------

# Conclusions and Next Steps

## Key Takeaways

**1. Strong downward trend (p \< 2e-16):** Shootings declined
consistently throughout the study period. This pre-existing trend
explains most of the reduction visible after January 2024.

**2. Weekend effect (IRR = 1.31, p \< 0.001):** About 31% more shootings
on Saturdays and Sundays, a pattern that remained stable before and
after the NFSU.

**3. Winter dip (IRR = 0.73, p \< 0.001):** Roughly 27% fewer shootings
in winter months. No other season differs significantly from fall once
trend is controlled.

**4. NFSU effect is not statistically significant (p = 0.947):** After
controlling for the pre-existing downward trend, the data cannot
distinguish the NFSU's contribution from trend continuation. This does
not mean the unit had no effect — it means the observational design
cannot isolate it from the background trend.

**5. No evidence of daily temporal clustering at this scale:** The
logged lag term (discrete self-excitation proxy) does not significantly
improve fit over M1, and the ACF/PACF of residuals are largely clean. If
Hawkes self-excitation is present, it likely operates at finer time
scales (hours) than daily bins can capture.

## What This Means for the Hawkes Model

**M1 is the estimated background intensity μ(t):**

$$\log\hat{\mu}(t) = 1.473 - 0.00127 \cdot \text{trend}_t + 0.273 \cdot \text{weekend}_t - 0.312 \cdot \mathbf{1}[\text{Winter}]$$

This function serves as the Poisson baseline into which the
self-excitation triggering kernel g(t − tᵢ) will be added in the full
Hawkes model. The clean ACF/PACF at the daily level is consistent with
the interpretation that self-excitation, if present, operates at
sub-daily resolution — which the continuous-time Hawkes process (rather
than this discretized GLM) is designed to capture.

**Next step:** Fit the spatio-temporal Hawkes process using the
`spatstat` or custom MLE approach, with μ(t) parameterized by these
covariate estimates as a fixed or estimated offset. \`\`\`
