---
title: "DASC 460: Spatial Analysis of Non-Fatal Shootings in Saint Paul — Grid-Level Modeling"
subtitle: "Incorporating Spatial Structure, Moran's I, LISA Hotspot Detection, and Weekly Poisson Regression"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

------------------------------------------------------------------------

# Setup

## Load Libraries

```{r warning=FALSE, message=FALSE}
library(rmarkdown)
library(knitr)
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(tidyr)
library(sf)
library(xml2)       # parse KML directly
library(spdep)      # Moran's I, LISA, spatial weights
library(MASS)       # negative binomial
library(lmtest)     # likelihood ratio test
library(AER)        # dispersion test
```

------------------------------------------------------------------------

# Section 1: Load and Prepare Data

## 1.1 Load Shooting Data

```{r}
file <- "Shootings_Full Data(new).xlsx"

full_data <- read_excel(file, sheet = "Full Data") %>%
  mutate(
    DATE       = as.Date(DATE),
    period     = if_else(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
    period     = factor(period, levels = c("Before NFSU", "After NFSU")),
    season     = case_when(
      month(DATE) %in% c(12, 1, 2)  ~ "Winter",
      month(DATE) %in% c(3, 4, 5)   ~ "Spring",
      month(DATE) %in% c(6, 7, 8)   ~ "Summer",
      month(DATE) %in% c(9, 10, 11) ~ "Fall"
    ),
    season     = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    week       = floor_date(DATE, "week", week_start = 1),
    week_index = as.integer(difftime(week, min(week, na.rm = TRUE), units = "weeks")),
    nfsu       = as.integer(DATE >= as.Date("2024-01-01")),
    GRID       = as.character(GRID)
  ) %>%
  filter(!is.na(Lat), !is.na(Long), !is.na(DATE), Lat != 0, Long != 0)

cat("Shooting records loaded:", nrow(full_data), "\n")
```

## 1.2 Load Police Grid from KMZ

We parse the KML directly using `xml2` to avoid geometry type issues that occur when `sf` reads KML MultiGeometry features.

```{r message=FALSE, warning=FALSE}
# Unzip the KMZ to get the KML file inside
kmz_path <- "StPaul_Police_Gri_LayerToKML.kmz"
unzip(kmz_path, exdir = tempdir(), overwrite = TRUE)
kml_path <- list.files(tempdir(), pattern = "\\.kml$", full.names = TRUE)[1]

# Parse KML with xml2
kml_doc    <- read_xml(kml_path)
ns         <- c(kml = "http://www.opengis.net/kml/2.2")
placemarks <- xml_find_all(kml_doc, ".//kml:Placemark", ns)

# Build sf object from scratch — one row per grid cell
features <- lapply(placemarks, function(pm) {
  # Grid ID is stored in the <name> tag
  grid_id  <- xml_text(xml_find_first(pm, "kml:name", ns))

  # Each Placemark may have one or more <Polygon> elements (grid 69 has 2)
  poly_nodes <- xml_find_all(
    pm, "kml:MultiGeometry/kml:Polygon/kml:outerBoundaryIs/kml:LinearRing/kml:coordinates", ns
  )
  # Also catch single Polygon (no MultiGeometry wrapper)
  if (length(poly_nodes) == 0) {
    poly_nodes <- xml_find_all(
      pm, "kml:Polygon/kml:outerBoundaryIs/kml:LinearRing/kml:coordinates", ns
    )
  }

  # Convert coordinate text to numeric matrix
  coords_list <- lapply(poly_nodes, function(co) {
    pts <- strsplit(trimws(xml_text(co)), "\\s+")[[1]]
    do.call(rbind, lapply(pts, function(p) {
      xy <- as.numeric(strsplit(p, ",")[[1]])
      c(xy[1], xy[2])  # lon, lat
    }))
  })

  # Build polygon or multipolygon geometry
  if (length(coords_list) == 1) {
    geom <- st_polygon(list(coords_list[[1]]))
  } else {
    geom <- st_multipolygon(lapply(coords_list, function(m) list(m)))
  }

  st_sf(Name = grid_id, geometry = st_sfc(geom, crs = 4326))
})

# Combine into one sf object
grid_sf <- do.call(rbind, features) %>%
  mutate(Name = as.character(Name))

cat("Grid cells loaded:", nrow(grid_sf), "\n")
cat("Geometry types:", paste(unique(st_geometry_type(grid_sf)), collapse = ", "), "\n")
cat("Sample grid IDs:", paste(head(grid_sf$Name, 8), collapse = ", "), "\n")
```

```{r}
# Preview the grid on a map
leaflet(grid_sf) %>%
  addTiles() %>%
  addPolygons(
    color       = "navy",
    weight      = 1,
    fillOpacity = 0.1,
    popup       = ~Name
  ) %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

## 1.3 Join Shootings to Grid Cells

```{r}
# Convert shooting points to sf object
shootings_sf <- full_data %>%
  filter(!is.na(Lat), !is.na(Long)) %>%
  st_as_sf(coords = c("Long", "Lat"), crs = 4326)

# Spatial join: assign each shooting to the grid cell it falls within
shootings_gridded <- st_join(shootings_sf, grid_sf, join = st_within) %>%
  rename(grid_id = Name) %>%
  st_drop_geometry()

cat("Matched:", sum(!is.na(shootings_gridded$grid_id)),
    "of", nrow(shootings_gridded), "shooting records to a grid\n")
cat("Grid cells with at least one shooting:",
    n_distinct(na.omit(shootings_gridded$grid_id)), "\n")
```

------------------------------------------------------------------------

# Section 2: Weekly Aggregation by Grid

Dr. Clark noted that daily aggregation produces too many zeros. Aggregating by week substantially reduces zero inflation and gives a cleaner panel for spatial and time-series modeling.

## 2.1 Build Weekly Grid-Level Panel

```{r}
all_weeks <- sort(unique(shootings_gridded$week))
all_grids <- sort(unique(grid_sf$Name))

# Complete panel: every week × grid combination
panel <- expand.grid(
  week    = all_weeks,
  grid_id = all_grids,
  stringsAsFactors = FALSE
) %>%
  as_tibble() %>%
  mutate(week = as.Date(week))

# Aggregate shootings by week × grid
weekly_counts <- shootings_gridded %>%
  filter(!is.na(grid_id), !is.na(week)) %>%
  group_by(week, grid_id) %>%
  summarize(count = n(), .groups = "drop")

# Join — missing combos become 0
panel <- panel %>%
  left_join(weekly_counts, by = c("week", "grid_id")) %>%
  replace_na(list(count = 0)) %>%
  mutate(
    nfsu       = as.integer(week >= as.Date("2024-01-01")),
    season     = case_when(
      month(week) %in% c(12, 1, 2)  ~ "Winter",
      month(week) %in% c(3, 4, 5)   ~ "Spring",
      month(week) %in% c(6, 7, 8)   ~ "Summer",
      month(week) %in% c(9, 10, 11) ~ "Fall"
    ),
    season     = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    week_index = as.integer(difftime(week, min(week), units = "weeks"))
  )

cat("Panel:", nrow(panel), "rows |",
    length(all_weeks), "weeks ×", length(all_grids), "grids\n")
```

## 2.2 Check Zero Proportion

```{r}
cat("Proportion of zeros (weekly grid-level):", round(mean(panel$count == 0), 3), "\n")

ggplot(panel, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Weekly Shooting Counts by Grid Cell",
    x     = "Shootings per Grid per Week",
    y     = "Frequency"
  ) +
  theme_minimal()
```

## 2.3 Identify High-Impact Grid Cells

```{r}
grid_totals <- panel %>%
  group_by(grid_id) %>%
  summarize(total = sum(count), .groups = "drop") %>%
  arrange(desc(total))

knitr::kable(head(grid_totals, 20), caption = "Top 20 Grids by Total Shootings")

# Top quartile of activity
top_grids <- grid_totals %>%
  filter(total >= quantile(grid_totals$total, 0.75)) %>%
  pull(grid_id)

cat("High-impact grids (top quartile):", length(top_grids), "\n")

panel_hot <- panel %>% filter(grid_id %in% top_grids)
cat("Zeros in high-impact panel:", round(mean(panel_hot$count == 0), 3), "\n")
```

## 2.4 Weekly Time Trend

```{r}
weekly_total <- panel %>%
  group_by(week, nfsu) %>%
  summarize(count = sum(count), .groups = "drop")

ggplot(weekly_total, aes(x = week, y = count)) +
  geom_line(color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "darkblue") +
  geom_vline(xintercept = as.Date("2024-01-01"), color = "red", linetype = "dashed") +
  annotate("text", x = as.Date("2024-01-01"), y = max(weekly_total$count),
           label = "NFSU Start", color = "red", hjust = -0.1, size = 3) +
  labs(title = "Weekly Shooting Counts Across All Grids", x = "Week", y = "Total Shootings") +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 3: Spatial Structure — Moran's I and LISA

## 3.1 Aggregate Spatially (Collapse Time)

```{r}
grid_spatial <- panel %>%
  group_by(grid_id) %>%
  summarize(total = sum(count), .groups = "drop")

grid_sf_counts <- grid_sf %>%
  rename(grid_id = Name) %>%
  left_join(grid_spatial, by = "grid_id") %>%
  replace_na(list(total = 0))
```

## 3.2 Map of Shooting Counts by Grid

```{r}
pal <- colorNumeric("YlOrRd", domain = grid_sf_counts$total)

leaflet(grid_sf_counts) %>%
  addTiles() %>%
  addPolygons(
    fillColor   = ~pal(total),
    fillOpacity = 0.7,
    color       = "black",
    weight      = 0.5,
    popup       = ~paste("<b>Grid:</b>", grid_id, "<br><b>Total Shootings:</b>", total)
  ) %>%
  addLegend(pal = pal, values = ~total, title = "Shootings", position = "bottomright")
```

## 3.3 Build Spatial Weights (Neighborhood Matrix)

```{r}
# Queen contiguity — grids sharing any edge or corner are neighbors
nb <- poly2nb(grid_sf_counts, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

cat("Grid regions:", length(nb), "\n")
cat("Average neighbors:", round(mean(card(nb)), 2), "\n")
```

## 3.4 Global Moran's I

```{r}
moran_result <- moran.test(
  grid_sf_counts$total,
  lw,
  zero.policy = TRUE,
  alternative = "greater"
)
print(moran_result)

moran.plot(
  grid_sf_counts$total, lw,
  zero.policy = TRUE,
  xlab = "Shooting Count (Grid)",
  ylab = "Spatially Lagged Count (Neighbors)",
  main = "Moran's I Scatter Plot — Total Shootings by Grid"
)
```

## 3.5 LISA — Local Indicators of Spatial Association

```{r}
lisa <- localmoran(grid_sf_counts$total, lw, zero.policy = TRUE, alternative = "two.sided")

grid_sf_counts <- grid_sf_counts %>%
  mutate(
    lisa_I   = lisa[, "Ii"],
    lisa_p   = lisa[, "Pr(z != E(Ii))"],
    z_score  = as.numeric(scale(total)),
    lag_z    = lag.listw(lw, z_score, zero.policy = TRUE),
    quad_sig = case_when(
      lisa_p >= 0.05             ~ "Not Significant",
      z_score > 0 & lag_z > 0   ~ "High-High (Hotspot)",
      z_score < 0 & lag_z < 0   ~ "Low-Low (Cold spot)",
      z_score > 0 & lag_z < 0   ~ "High-Low (Outlier)",
      z_score < 0 & lag_z > 0   ~ "Low-High (Outlier)"
    ),
    quad_sig = factor(quad_sig, levels = c(
      "High-High (Hotspot)", "Low-Low (Cold spot)",
      "High-Low (Outlier)", "Low-High (Outlier)", "Not Significant"
    ))
  )
```

## 3.6 LISA Cluster Map

```{r}
lisa_colors <- c(
  "High-High (Hotspot)" = "#d7191c",
  "Low-Low (Cold spot)" = "#2c7bb6",
  "High-Low (Outlier)"  = "#fdae61",
  "Low-High (Outlier)"  = "#abd9e9",
  "Not Significant"     = "#f0f0f0"
)

pal_lisa <- colorFactor(palette = lisa_colors, domain = names(lisa_colors))

leaflet(grid_sf_counts) %>%
  addTiles() %>%
  addPolygons(
    fillColor   = ~pal_lisa(quad_sig),
    fillOpacity = 0.75,
    color       = "black",
    weight      = 0.5,
    popup       = ~paste(
      "<b>Grid:</b>",    grid_id, "<br>",
      "<b>Total:</b>",   total,   "<br>",
      "<b>LISA:</b>",    quad_sig, "<br>",
      "<b>p-value:</b>", round(lisa_p, 4)
    )
  ) %>%
  addLegend(pal = pal_lisa, values = ~quad_sig,
            title = "LISA Cluster Type", position = "bottomright") %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

## 3.7 LISA Summary Table

```{r}
grid_sf_counts %>%
  st_drop_geometry() %>%
  group_by(quad_sig) %>%
  summarize(n_grids = n(), .groups = "drop") %>%
  arrange(desc(n_grids)) %>%
  knitr::kable(caption = "LISA Cluster Classification — Grid Count")
```

------------------------------------------------------------------------

# Section 4: Poisson Regression (Weekly, Time-Level)

## 4.1 Add Lag Variable

```{r}
panel <- panel %>%
  arrange(grid_id, week) %>%
  group_by(grid_id) %>%
  mutate(
    lag_count     = lag(count, 1),
    log_lag_count = log(lag_count + 1)
  ) %>%
  ungroup() %>%
  filter(!is.na(lag_count))
```

## 4.2 Weekly Aggregated Models (Time Only)

```{r}
weekly_model_data <- panel %>%
  group_by(week, week_index, nfsu, season) %>%
  summarize(
    count     = sum(count),
    lag_count = sum(lag_count),
    .groups   = "drop"
  ) %>%
  mutate(log_lag = log(lag_count + 1))

# Four nested models
m1 <- glm(count ~ week_index,
           data = weekly_model_data, family = poisson(link = "log"))

m2 <- glm(count ~ week_index + nfsu,
           data = weekly_model_data, family = poisson(link = "log"))

m3 <- glm(count ~ week_index + nfsu + season,
           data = weekly_model_data, family = poisson(link = "log"))

m4 <- glm(count ~ week_index + nfsu + season + log_lag,
           data = weekly_model_data, family = poisson(link = "log"))

# Comparison table
data.frame(
  Model     = c("M1: Trend", "M2: +NFSU", "M3: +Season", "M4: +Log-Lag"),
  AIC       = round(c(AIC(m1), AIC(m2), AIC(m3), AIC(m4)), 2),
  BIC       = round(c(BIC(m1), BIC(m2), BIC(m3), BIC(m4)), 2),
  Resid_Dev = round(c(deviance(m1), deviance(m2), deviance(m3), deviance(m4)), 2)
) %>%
  knitr::kable(caption = "Model Comparison — AIC / BIC / Residual Deviance")
```

## 4.3 Likelihood Ratio Tests

```{r}
lrtest(m1, m2)   # Is NFSU significant?
lrtest(m2, m3)   # Does season improve fit?
lrtest(m3, m4)   # Does log-lag improve fit?
```

## 4.4 Dispersion Check

```{r}
dispersiontest(m4)

cat("Dispersion (deviance/df):", round(deviance(m4) / df.residual(m4), 3), "\n")

# Fit Negative Binomial for comparison
m4_nb <- glm.nb(count ~ week_index + nfsu + season + log_lag,
                data = weekly_model_data)

cat("Poisson AIC:", round(AIC(m4),    2), "\n")
cat("NegBin  AIC:", round(AIC(m4_nb), 2), "\n")
cat("Use whichever has the lower AIC.\n")
```

## 4.5 Best Model Summary and IRRs

```{r}
# Update 'best_model' to m4_nb if NegBin AIC is lower
best_model <- m4

summary(best_model)

# Exponentiated coefficients = Incidence Rate Ratios
irr <- exp(cbind(IRR = coef(best_model), confint(best_model)))
knitr::kable(round(irr, 4), caption = "Incidence Rate Ratios")
```

------------------------------------------------------------------------

# Section 5: Model Diagnostics

## 5.1 Deviance Residuals vs Fitted

```{r}
weekly_model_data <- weekly_model_data %>%
  mutate(
    fitted_vals = fitted(best_model),
    dev_resid   = residuals(best_model, type = "deviance")
  )

ggplot(weekly_model_data, aes(x = fitted_vals, y = dev_resid)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "darkblue") +
  labs(title = "Deviance Residuals vs Fitted Values",
       x = "Fitted Values", y = "Deviance Residuals") +
  theme_minimal()
```

## 5.2 ACF and PACF of Residuals

```{r fig.height=4}
par(mfrow = c(1, 2))
acf(weekly_model_data$dev_resid,
    main = "ACF — Deviance Residuals", lag.max = 30)
pacf(weekly_model_data$dev_resid,
     main = "PACF — Deviance Residuals", lag.max = 30)
par(mfrow = c(1, 1))
```

## 5.3 QQ Plot

```{r}
ggplot(weekly_model_data, aes(sample = dev_resid)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot — Deviance Residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```

## 5.4 Residuals Over Time

```{r}
ggplot(weekly_model_data, aes(x = week, y = dev_resid)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), color = "darkred", linetype = "dotted") +
  annotate("text", x = as.Date("2024-01-01"),
           y = max(weekly_model_data$dev_resid, na.rm = TRUE),
           label = "NFSU", color = "darkred", hjust = -0.1, size = 3) +
  labs(title = "Deviance Residuals Over Time",
       x = "Week", y = "Deviance Residual") +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 6: Spatial Residual Diagnostics

## 6.1 Grid-Level Poisson Model

```{r}
panel_model <- panel %>%
  filter(grid_id %in% top_grids) %>%
  mutate(grid_id = factor(grid_id))

m_grid <- glm(
  count ~ week_index + nfsu + season + log_lag_count + grid_id,
  data   = panel_model,
  family = poisson(link = "log")
)

cat("Grid-level Poisson AIC:", round(AIC(m_grid), 2), "\n")
cat("Dispersion:", round(deviance(m_grid) / df.residual(m_grid), 3), "\n")
```

## 6.2 Spatial Residual Map

```{r}
panel_model <- panel_model %>%
  mutate(dev_resid = residuals(m_grid, type = "deviance"))

grid_resid <- panel_model %>%
  group_by(grid_id) %>%
  summarize(mean_resid = mean(dev_resid), .groups = "drop") %>%
  mutate(grid_id = as.character(grid_id))

grid_sf_resid <- grid_sf %>%
  rename(grid_id = Name) %>%
  inner_join(grid_resid, by = "grid_id")

pal_resid <- colorNumeric(c("blue", "white", "red"), domain = grid_sf_resid$mean_resid)

leaflet(grid_sf_resid) %>%
  addTiles() %>%
  addPolygons(
    fillColor   = ~pal_resid(mean_resid),
    fillOpacity = 0.7,
    color       = "black", weight = 0.5,
    popup = ~paste("<b>Grid:</b>", grid_id,
                   "<br><b>Mean Residual:</b>", round(mean_resid, 3))
  ) %>%
  addLegend(pal = pal_resid, values = ~mean_resid,
            title = "Mean Dev. Residual<br>(Red = underpredicted)",
            position = "bottomright")
```

## 6.3 Moran's I on Spatial Residuals

```{r}
grid_sf_moran <- grid_sf %>%
  rename(grid_id = Name) %>%
  left_join(grid_resid, by = "grid_id") %>%
  replace_na(list(mean_resid = 0))

nb_resid <- poly2nb(grid_sf_moran, queen = TRUE)
lw_resid <- nb2listw(nb_resid, style = "W", zero.policy = TRUE)

moran_resid <- moran.test(grid_sf_moran$mean_resid, lw_resid, zero.policy = TRUE)
print(moran_resid)
cat("\nIf significant: spatial structure remains — space-time model needed.\n")
```

------------------------------------------------------------------------

# Section 7: Before vs After NFSU — Spatial Comparison

## 7.1 LISA Before vs After

```{r}
run_lisa <- function(df_period, grid_polygon, nb_weights) {
  counts <- df_period %>%
    group_by(grid_id) %>%
    summarize(total = sum(count), .groups = "drop")

  gdf <- grid_polygon %>%
    left_join(counts, by = "grid_id") %>%
    replace_na(list(total = 0))

  lisa_out <- localmoran(gdf$total, nb_weights, zero.policy = TRUE)

  gdf %>%
    mutate(
      lisa_p   = lisa_out[, "Pr(z != E(Ii))"],
      z_score  = as.numeric(scale(total)),
      lag_z    = lag.listw(nb_weights, z_score, zero.policy = TRUE),
      quad_sig = case_when(
        lisa_p >= 0.05             ~ "Not Significant",
        z_score > 0 & lag_z > 0   ~ "High-High (Hotspot)",
        z_score < 0 & lag_z < 0   ~ "Low-Low (Cold spot)",
        z_score > 0 & lag_z < 0   ~ "High-Low (Outlier)",
        z_score < 0 & lag_z > 0   ~ "Low-High (Outlier)"
      ),
      quad_sig = factor(quad_sig, levels = c(
        "High-High (Hotspot)", "Low-Low (Cold spot)",
        "High-Low (Outlier)", "Low-High (Outlier)", "Not Significant"
      ))
    )
}

grid_base    <- grid_sf %>% rename(grid_id = Name)
panel_before <- panel %>% filter(nfsu == 0)
panel_after  <- panel %>% filter(nfsu == 1)

lisa_before <- run_lisa(panel_before, grid_base, lw)
lisa_after  <- run_lisa(panel_after,  grid_base, lw)
```

```{r}
pal_lisa2 <- colorFactor(
  palette = c("#d7191c", "#2c7bb6", "#fdae61", "#abd9e9", "#f0f0f0"),
  levels  = c("High-High (Hotspot)", "Low-Low (Cold spot)",
              "High-Low (Outlier)", "Low-High (Outlier)", "Not Significant")
)

# Before NFSU
leaflet(lisa_before) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal_lisa2(quad_sig), fillOpacity = 0.75,
    color = "black", weight = 0.5,
    popup = ~paste("<b>Grid:</b>", grid_id, "<br><b>LISA:</b>", quad_sig)
  ) %>%
  addLegend(pal = pal_lisa2, values = ~quad_sig,
            title = "LISA: Before NFSU", position = "bottomright") %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

```{r}
# After NFSU
leaflet(lisa_after) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal_lisa2(quad_sig), fillOpacity = 0.75,
    color = "black", weight = 0.5,
    popup = ~paste("<b>Grid:</b>", grid_id, "<br><b>LISA:</b>", quad_sig)
  ) %>%
  addLegend(pal = pal_lisa2, values = ~quad_sig,
            title = "LISA: After NFSU", position = "bottomright") %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

## 7.2 Hotspot Stability Table

```{r}
hotspot_comparison <- lisa_before %>%
  st_drop_geometry() %>%
  dplyr::select(grid_id, before_quad = quad_sig) %>%
  left_join(
    lisa_after %>% st_drop_geometry() %>% dplyr::select(grid_id, after_quad = quad_sig),
    by = "grid_id"
  ) %>%
  mutate(
    status = case_when(
      before_quad == "High-High (Hotspot)" & after_quad == "High-High (Hotspot)" ~ "Persistent Hotspot",
      before_quad == "High-High (Hotspot)" & after_quad != "High-High (Hotspot)" ~ "Cooled (NFSU effect?)",
      before_quad != "High-High (Hotspot)" & after_quad == "High-High (Hotspot)" ~ "Emerging Hotspot",
      TRUE ~ "Neither"
    )
  )

hotspot_comparison %>%
  filter(status != "Neither") %>%
  arrange(status, grid_id) %>%
  knitr::kable(caption = "Hotspot Status Before vs After NFSU")
```

------------------------------------------------------------------------

# Section 8: Session Info

```{r}
sessionInfo()
```
