---
title: "DASC 460: Spatial Analysis of Non-Fatal Shootings in Saint Paul — Grid-Level Modeling"
subtitle: "Incorporating Spatial Structure, Moran's I, LISA Hotspot Detection, and Weekly Poisson Regression"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

# Setup

## Load Libraries

```{r warning=FALSE, message=FALSE}
library(rmarkdown)
library(knitr)
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(tidyr)
library(sf)
library(xml2)       # parse KML directly
library(spdep)      # Moran's I, LISA, spatial weights
library(MASS)       # negative binomial
library(lmtest)     # likelihood ratio test
library(AER)        # dispersion test
```

------------------------------------------------------------------------

# Section 1: Load and Prepare Data

## 1.1 Load Shooting Data

```{r}
file <- "Shootings_Full Data(new).xlsx"

full_data <- read_excel(file, sheet = "Full Data") %>%
  mutate(
    DATE       = as.Date(DATE),
    period     = if_else(DATE < as.Date("2024-01-01"), "Before NFSU", "After NFSU"),
    period     = factor(period, levels = c("Before NFSU", "After NFSU")),
    season     = case_when(
      month(DATE) %in% c(12, 1, 2)  ~ "Winter",
      month(DATE) %in% c(3, 4, 5)   ~ "Spring",
      month(DATE) %in% c(6, 7, 8)   ~ "Summer",
      month(DATE) %in% c(9, 10, 11) ~ "Fall"
    ),
    season     = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    week       = floor_date(DATE, "week", week_start = 1),
    week_index = as.integer(difftime(week, min(week, na.rm = TRUE), units = "weeks")),
    nfsu       = as.integer(DATE >= as.Date("2024-01-01")),
    GRID       = as.character(GRID)
  ) %>%
  filter(!is.na(Lat), !is.na(Long), !is.na(DATE), Lat != 0, Long != 0)

cat("Shooting records loaded:", nrow(full_data), "\n")
```

## 1.2 Load Police Grid from KMZ

We parse the KML directly using `xml2` to avoid geometry type issues
that occur when `sf` reads KML MultiGeometry features.

```{r message=FALSE, warning=FALSE}
# Unzip the KMZ to get the KML file inside
kmz_path <- "StPaul_Police_Gri_LayerToKML.kmz"
unzip(kmz_path, exdir = tempdir(), overwrite = TRUE)
kml_path <- list.files(tempdir(), pattern = "\\.kml$", full.names = TRUE)[1]

# Parse KML with xml2
kml_doc    <- read_xml(kml_path)
ns         <- c(kml = "http://www.opengis.net/kml/2.2")
placemarks <- xml_find_all(kml_doc, ".//kml:Placemark", ns)

# Build sf object from scratch — one row per grid cell
features <- lapply(placemarks, function(pm) {
  # Grid ID is stored in the <name> tag
  grid_id  <- xml_text(xml_find_first(pm, "kml:name", ns))

  # Each Placemark may have one or more <Polygon> elements (grid 69 has 2)
  poly_nodes <- xml_find_all(
    pm, "kml:MultiGeometry/kml:Polygon/kml:outerBoundaryIs/kml:LinearRing/kml:coordinates", ns
  )
  # Also catch single Polygon (no MultiGeometry wrapper)
  if (length(poly_nodes) == 0) {
    poly_nodes <- xml_find_all(
      pm, "kml:Polygon/kml:outerBoundaryIs/kml:LinearRing/kml:coordinates", ns
    )
  }

  # Convert coordinate text to numeric matrix
  coords_list <- lapply(poly_nodes, function(co) {
    pts <- strsplit(trimws(xml_text(co)), "\\s+")[[1]]
    do.call(rbind, lapply(pts, function(p) {
      xy <- as.numeric(strsplit(p, ",")[[1]])
      c(xy[1], xy[2])  # lon, lat
    }))
  })

  # Build polygon or multipolygon geometry
  if (length(coords_list) == 1) {
    geom <- st_polygon(list(coords_list[[1]]))
  } else {
    geom <- st_multipolygon(lapply(coords_list, function(m) list(m)))
  }

  st_sf(Name = grid_id, geometry = st_sfc(geom, crs = 4326))
})

# Combine into one sf object
grid_sf <- do.call(rbind, features) %>%
  mutate(Name = as.character(Name))

cat("Grid cells loaded:", nrow(grid_sf), "\n")
cat("Geometry types:", paste(unique(st_geometry_type(grid_sf)), collapse = ", "), "\n")
cat("Sample grid IDs:", paste(head(grid_sf$Name, 8), collapse = ", "), "\n")
```

```{r}
# Preview the grid on a map
leaflet(grid_sf) %>%
  addTiles() %>%
  addPolygons(
    color       = "navy",
    weight      = 1,
    fillOpacity = 0.1,
    popup       = ~Name
  ) %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

## 1.3 Join Shootings to Grid Cells

```{r}
# Convert shooting points to sf object
shootings_sf <- full_data %>%
  filter(!is.na(Lat), !is.na(Long)) %>%
  st_as_sf(coords = c("Long", "Lat"), crs = 4326)

# Spatial join: assign each shooting to the grid cell it falls within
shootings_gridded <- st_join(shootings_sf, grid_sf, join = st_within) %>%
  rename(grid_id = Name) %>%
  st_drop_geometry()

cat("Matched:", sum(!is.na(shootings_gridded$grid_id)),
    "of", nrow(shootings_gridded), "shooting records to a grid\n")
cat("Grid cells with at least one shooting:",
    n_distinct(na.omit(shootings_gridded$grid_id)), "\n")
```

------------------------------------------------------------------------

# Section 2: Weekly Aggregation by Grid

Dr. Clark noted that daily aggregation produces too many zeros.
Aggregating by week substantially reduces zero inflation and gives a
cleaner panel for spatial and time-series modeling.

## 2.1 Build Weekly Grid-Level Panel

```{r}
all_weeks <- sort(unique(shootings_gridded$week))
all_grids <- sort(unique(grid_sf$Name))

# Complete panel: every week × grid combination
panel <- expand.grid(
  week    = all_weeks,
  grid_id = all_grids,
  stringsAsFactors = FALSE
) %>%
  as_tibble() %>%
  mutate(week = as.Date(week))

# Aggregate shootings by week × grid
weekly_counts <- shootings_gridded %>%
  filter(!is.na(grid_id), !is.na(week)) %>%
  group_by(week, grid_id) %>%
  summarize(count = n(), .groups = "drop")

# Join — missing combos become 0
panel <- panel %>%
  left_join(weekly_counts, by = c("week", "grid_id")) %>%
  replace_na(list(count = 0)) %>%
  mutate(
    nfsu       = as.integer(week >= as.Date("2024-01-01")),
    season     = case_when(
      month(week) %in% c(12, 1, 2)  ~ "Winter",
      month(week) %in% c(3, 4, 5)   ~ "Spring",
      month(week) %in% c(6, 7, 8)   ~ "Summer",
      month(week) %in% c(9, 10, 11) ~ "Fall"
    ),
    season     = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    week_index = as.integer(difftime(week, min(week), units = "weeks"))
  )

cat("Panel:", nrow(panel), "rows |",
    length(all_weeks), "weeks ×", length(all_grids), "grids\n")
```

## 2.2 Check Zero Proportion

```{r}
cat("Proportion of zeros (weekly grid-level):", round(mean(panel$count == 0), 3), "\n")

ggplot(panel, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Weekly Shooting Counts by Grid Cell",
    x     = "Shootings per Grid per Week",
    y     = "Frequency"
  ) +
  theme_minimal()
```

## 2.3 Identify High-Impact Grid Cells

```{r}
grid_totals <- panel %>%
  group_by(grid_id) %>%
  summarize(total = sum(count), .groups = "drop") %>%
  arrange(desc(total))

knitr::kable(head(grid_totals, 20), caption = "Top 20 Grids by Total Shootings")

# Top quartile of activity
top_grids <- grid_totals %>%
  filter(total >= quantile(grid_totals$total, 0.75)) %>%
  pull(grid_id)

cat("High-impact grids (top quartile):", length(top_grids), "\n")

panel_hot <- panel %>% filter(grid_id %in% top_grids)
cat("Zeros in high-impact panel:", round(mean(panel_hot$count == 0), 3), "\n")
```

## 2.4 Weekly Time Trend

```{r}
weekly_total <- panel %>%
  group_by(week, nfsu) %>%
  summarize(count = sum(count), .groups = "drop")

ggplot(weekly_total, aes(x = week, y = count)) +
  geom_line(color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "darkblue") +
  geom_vline(xintercept = as.Date("2024-01-01"), color = "red", linetype = "dashed") +
  annotate("text", x = as.Date("2024-01-01"), y = max(weekly_total$count),
           label = "NFSU Start", color = "red", hjust = -0.1, size = 3) +
  labs(title = "Weekly Shooting Counts Across All Grids", x = "Week", y = "Total Shootings") +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 3: Spatial Structure — Moran's I and LISA

## 3.1 Aggregate Spatially (Collapse Time)

```{r}
grid_spatial <- panel %>%
  group_by(grid_id) %>%
  summarize(total = sum(count), .groups = "drop")

grid_sf_counts <- grid_sf %>%
  rename(grid_id = Name) %>%
  left_join(grid_spatial, by = "grid_id") %>%
  replace_na(list(total = 0))
```

## 3.2 Map of Shooting Counts by Grid

```{r}
pal <- colorNumeric("YlOrRd", domain = grid_sf_counts$total)

leaflet(grid_sf_counts) %>%
  addTiles() %>%
  addPolygons(
    fillColor   = ~pal(total),
    fillOpacity = 0.7,
    color       = "black",
    weight      = 0.5,
    popup       = ~paste("<b>Grid:</b>", grid_id, "<br><b>Total Shootings:</b>", total)
  ) %>%
  addLegend(pal = pal, values = ~total, title = "Shootings", position = "bottomright")
```

## 3.3 Build Spatial Weights (Neighborhood Matrix)

```{r}
# Queen contiguity — grids sharing any edge or corner are neighbors
nb <- poly2nb(grid_sf_counts, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

cat("Grid regions:", length(nb), "\n")
cat("Average neighbors:", round(mean(card(nb)), 2), "\n")
```

## 3.4 Global Moran's I

```{r}
moran_result <- moran.test(
  grid_sf_counts$total,
  lw,
  zero.policy = TRUE,
  alternative = "greater"
)
print(moran_result)

moran.plot(
  grid_sf_counts$total, lw,
  zero.policy = TRUE,
  xlab = "Shooting Count (Grid)",
  ylab = "Spatially Lagged Count (Neighbors)",
  main = "Moran's I Scatter Plot — Total Shootings by Grid"
)
```

## 3.5 LISA — Local Indicators of Spatial Association

```{r}
lisa <- localmoran(grid_sf_counts$total, lw, zero.policy = TRUE, alternative = "two.sided")

grid_sf_counts <- grid_sf_counts %>%
  mutate(
    lisa_I   = lisa[, "Ii"],
    lisa_p   = lisa[, "Pr(z != E(Ii))"],
    z_score  = as.numeric(scale(total)),
    lag_z    = lag.listw(lw, z_score, zero.policy = TRUE),
    quad_sig = case_when(
      lisa_p >= 0.05             ~ "Not Significant",
      z_score > 0 & lag_z > 0   ~ "High-High (Hotspot)",
      z_score < 0 & lag_z < 0   ~ "Low-Low (Cold spot)",
      z_score > 0 & lag_z < 0   ~ "High-Low (Outlier)",
      z_score < 0 & lag_z > 0   ~ "Low-High (Outlier)"
    ),
    quad_sig = factor(quad_sig, levels = c(
      "High-High (Hotspot)", "Low-Low (Cold spot)",
      "High-Low (Outlier)", "Low-High (Outlier)", "Not Significant"
    ))
  )
```

## 3.6 LISA Cluster Map

```{r}
lisa_colors <- c(
  "High-High (Hotspot)" = "#d7191c",
  "Low-Low (Cold spot)" = "#2c7bb6",
  "High-Low (Outlier)"  = "#fdae61",
  "Low-High (Outlier)"  = "#abd9e9",
  "Not Significant"     = "#f0f0f0"
)

pal_lisa <- colorFactor(palette = lisa_colors, domain = names(lisa_colors))

leaflet(grid_sf_counts) %>%
  addTiles() %>%
  addPolygons(
    fillColor   = ~pal_lisa(quad_sig),
    fillOpacity = 0.75,
    color       = "black",
    weight      = 0.5,
    popup       = ~paste(
      "<b>Grid:</b>",    grid_id, "<br>",
      "<b>Total:</b>",   total,   "<br>",
      "<b>LISA:</b>",    quad_sig, "<br>",
      "<b>p-value:</b>", round(lisa_p, 4)
    )
  ) %>%
  addLegend(pal = pal_lisa, values = ~quad_sig,
            title = "LISA Cluster Type", position = "bottomright") %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

## 3.7 LISA Summary Table

```{r}
grid_sf_counts %>%
  st_drop_geometry() %>%
  group_by(quad_sig) %>%
  summarize(n_grids = n(), .groups = "drop") %>%
  arrange(desc(n_grids)) %>%
  knitr::kable(caption = "LISA Cluster Classification — Grid Count")
```

------------------------------------------------------------------------

# Section 4: Poisson Regression (Weekly, Time-Level)

## 4.1 Add Lag Variable

```{r}
panel <- panel %>%
  arrange(grid_id, week) %>%
  group_by(grid_id) %>%
  mutate(
    lag_count     = lag(count, 1),
    log_lag_count = log(lag_count + 1)
  ) %>%
  ungroup() %>%
  filter(!is.na(lag_count))
```

## 4.2 Weekly Aggregated Models (Time Only)

```{r}
weekly_model_data <- panel %>%
  group_by(week, week_index, nfsu, season) %>%
  summarize(
    count     = sum(count),
    lag_count = sum(lag_count),
    .groups   = "drop"
  ) %>%
  mutate(log_lag = log(lag_count + 1))

# Four nested models
m1 <- glm(count ~ week_index,
           data = weekly_model_data, family = poisson(link = "log"))

m2 <- glm(count ~ week_index + nfsu,
           data = weekly_model_data, family = poisson(link = "log"))

m3 <- glm(count ~ week_index + nfsu + season,
           data = weekly_model_data, family = poisson(link = "log"))

m4 <- glm(count ~ week_index + nfsu + season + log_lag,
           data = weekly_model_data, family = poisson(link = "log"))

# Comparison table
data.frame(
  Model     = c("M1: Trend", "M2: +NFSU", "M3: +Season", "M4: +Log-Lag"),
  AIC       = round(c(AIC(m1), AIC(m2), AIC(m3), AIC(m4)), 2),
  BIC       = round(c(BIC(m1), BIC(m2), BIC(m3), BIC(m4)), 2),
  Resid_Dev = round(c(deviance(m1), deviance(m2), deviance(m3), deviance(m4)), 2)
) %>%
  knitr::kable(caption = "Model Comparison — AIC / BIC / Residual Deviance")
```

## 4.3 Likelihood Ratio Tests

```{r}
lrtest(m1, m2)   # Is NFSU significant?
lrtest(m2, m3)   # Does season improve fit?
lrtest(m3, m4)   # Does log-lag improve fit?
```

## 4.4 Dispersion Check

```{r}
dispersiontest(m4)

cat("Dispersion (deviance/df):", round(deviance(m4) / df.residual(m4), 3), "\n")

# Fit Negative Binomial for comparison
m4_nb <- glm.nb(count ~ week_index + nfsu + season + log_lag,
                data = weekly_model_data)

cat("Poisson AIC:", round(AIC(m4),    2), "\n")
cat("NegBin  AIC:", round(AIC(m4_nb), 2), "\n")
cat("Use whichever has the lower AIC.\n")
```

## 4.5 Best Model Summary and IRRs

```{r}
# Update 'best_model' to m4_nb if NegBin AIC is lower
best_model <- m4

summary(best_model)

# Exponentiated coefficients = Incidence Rate Ratios
irr <- exp(cbind(IRR = coef(best_model), confint(best_model)))
knitr::kable(round(irr, 4), caption = "Incidence Rate Ratios")
```

------------------------------------------------------------------------

# Section 5: Model Diagnostics

## 5.1 Deviance Residuals vs Fitted

```{r}
weekly_model_data <- weekly_model_data %>%
  mutate(
    fitted_vals = fitted(best_model),
    dev_resid   = residuals(best_model, type = "deviance")
  )

ggplot(weekly_model_data, aes(x = fitted_vals, y = dev_resid)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "darkblue") +
  labs(title = "Deviance Residuals vs Fitted Values",
       x = "Fitted Values", y = "Deviance Residuals") +
  theme_minimal()
```

## 5.2 ACF and PACF of Residuals

```{r fig.height=4}
op <- par(no.readonly = TRUE)
on.exit(par(op), add = TRUE)
par(mfrow = c(1, 2))
acf(weekly_model_data$dev_resid,
    main = "ACF — Deviance Residuals", lag.max = 30)
pacf(weekly_model_data$dev_resid,
     main = "PACF — Deviance Residuals", lag.max = 30)
par(op)
```

## 5.3 QQ Plot

```{r}
ggplot(weekly_model_data, aes(sample = dev_resid)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot — Deviance Residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```

## 5.4 Residuals Over Time

```{r}
ggplot(weekly_model_data, aes(x = week, y = dev_resid)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), color = "darkred", linetype = "dotted") +
  annotate("text", x = as.Date("2024-01-01"),
           y = max(weekly_model_data$dev_resid, na.rm = TRUE),
           label = "NFSU", color = "darkred", hjust = -0.1, size = 3) +
  labs(title = "Deviance Residuals Over Time",
       x = "Week", y = "Deviance Residual") +
  theme_minimal()
```

------------------------------------------------------------------------

# Section 6: Spatial Residual Diagnostics

## 6.1 Grid-Level Poisson Model

```{r}
panel_model <- panel %>%
  filter(grid_id %in% top_grids) %>%
  mutate(grid_id = factor(grid_id))

m_grid <- glm(
  count ~ week_index + nfsu + season + log_lag_count + grid_id,
  data   = panel_model,
  family = poisson(link = "log")
)

cat("Grid-level Poisson AIC:", round(AIC(m_grid), 2), "\n")
cat("Dispersion:", round(deviance(m_grid) / df.residual(m_grid), 3), "\n")
```

## 6.2 Spatial Residual Map

```{r}
panel_model <- panel_model %>%
  mutate(dev_resid = residuals(m_grid, type = "deviance"))

grid_resid <- panel_model %>%
  group_by(grid_id) %>%
  summarize(mean_resid = mean(dev_resid), .groups = "drop") %>%
  mutate(grid_id = as.character(grid_id))

grid_sf_resid <- grid_sf %>%
  rename(grid_id = Name) %>%
  inner_join(grid_resid, by = "grid_id")

pal_resid <- colorNumeric(c("blue", "white", "red"), domain = grid_sf_resid$mean_resid)

leaflet(grid_sf_resid) %>%
  addTiles() %>%
  addPolygons(
    fillColor   = ~pal_resid(mean_resid),
    fillOpacity = 0.7,
    color       = "black", weight = 0.5,
    popup = ~paste("<b>Grid:</b>", grid_id,
                   "<br><b>Mean Residual:</b>", round(mean_resid, 3))
  ) %>%
  addLegend(pal = pal_resid, values = ~mean_resid,
            title = "Mean Dev. Residual<br>(Red = underpredicted)",
            position = "bottomright")
```

## 6.3 Moran's I on Spatial Residuals

```{r}
grid_sf_moran <- grid_sf %>%
  rename(grid_id = Name) %>%
  left_join(grid_resid, by = "grid_id") %>%
  replace_na(list(mean_resid = 0))

nb_resid <- poly2nb(grid_sf_moran, queen = TRUE)
lw_resid <- nb2listw(nb_resid, style = "W", zero.policy = TRUE)

moran_resid <- moran.test(grid_sf_moran$mean_resid, lw_resid, zero.policy = TRUE)
print(moran_resid)
cat("\nIf significant: spatial structure remains — space-time model needed.\n")
```

------------------------------------------------------------------------

# Section 7: Before vs After NFSU — Spatial Comparison

## 7.1 LISA Before vs After

```{r}
run_lisa <- function(df_period, grid_polygon, nb_weights) {
  counts <- df_period %>%
    group_by(grid_id) %>%
    summarize(total = sum(count), .groups = "drop")

  gdf <- grid_polygon %>%
    left_join(counts, by = "grid_id") %>%
    replace_na(list(total = 0))

  lisa_out <- localmoran(gdf$total, nb_weights, zero.policy = TRUE)

  gdf %>%
    mutate(
      lisa_p   = lisa_out[, "Pr(z != E(Ii))"],
      z_score  = as.numeric(scale(total)),
      lag_z    = lag.listw(nb_weights, z_score, zero.policy = TRUE),
      quad_sig = case_when(
        lisa_p >= 0.05             ~ "Not Significant",
        z_score > 0 & lag_z > 0   ~ "High-High (Hotspot)",
        z_score < 0 & lag_z < 0   ~ "Low-Low (Cold spot)",
        z_score > 0 & lag_z < 0   ~ "High-Low (Outlier)",
        z_score < 0 & lag_z > 0   ~ "Low-High (Outlier)"
      ),
      quad_sig = factor(quad_sig, levels = c(
        "High-High (Hotspot)", "Low-Low (Cold spot)",
        "High-Low (Outlier)", "Low-High (Outlier)", "Not Significant"
      ))
    )
}

grid_base    <- grid_sf %>% rename(grid_id = Name)
panel_before <- panel %>% filter(nfsu == 0)
panel_after  <- panel %>% filter(nfsu == 1)

lisa_before <- run_lisa(panel_before, grid_base, lw)
lisa_after  <- run_lisa(panel_after,  grid_base, lw)
```

```{r}
pal_lisa2 <- colorFactor(
  palette = c("#d7191c", "#2c7bb6", "#fdae61", "#abd9e9", "#f0f0f0"),
  levels  = c("High-High (Hotspot)", "Low-Low (Cold spot)",
              "High-Low (Outlier)", "Low-High (Outlier)", "Not Significant")
)

# Before NFSU
leaflet(lisa_before) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal_lisa2(quad_sig), fillOpacity = 0.75,
    color = "black", weight = 0.5,
    popup = ~paste("<b>Grid:</b>", grid_id, "<br><b>LISA:</b>", quad_sig)
  ) %>%
  addLegend(pal = pal_lisa2, values = ~quad_sig,
            title = "LISA: Before NFSU", position = "bottomright") %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

```{r}
# After NFSU
leaflet(lisa_after) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal_lisa2(quad_sig), fillOpacity = 0.75,
    color = "black", weight = 0.5,
    popup = ~paste("<b>Grid:</b>", grid_id, "<br><b>LISA:</b>", quad_sig)
  ) %>%
  addLegend(pal = pal_lisa2, values = ~quad_sig,
            title = "LISA: After NFSU", position = "bottomright") %>%
  setView(lng = -93.09, lat = 44.95, zoom = 12)
```

## 7.2 Hotspot Stability Table

```{r}
hotspot_comparison <- lisa_before %>%
  st_drop_geometry() %>%
  dplyr::select(grid_id, before_quad = quad_sig) %>%
  left_join(
    lisa_after %>% st_drop_geometry() %>% dplyr::select(grid_id, after_quad = quad_sig),
    by = "grid_id"
  ) %>%
  mutate(
    status = case_when(
      before_quad == "High-High (Hotspot)" & after_quad == "High-High (Hotspot)" ~ "Persistent Hotspot",
      before_quad == "High-High (Hotspot)" & after_quad != "High-High (Hotspot)" ~ "Cooled (NFSU effect?)",
      before_quad != "High-High (Hotspot)" & after_quad == "High-High (Hotspot)" ~ "Emerging Hotspot",
      TRUE ~ "Neither"
    )
  )

hotspot_comparison %>%
  filter(status != "Neither") %>%
  arrange(status, grid_id) %>%
  knitr::kable(caption = "Hotspot Status Before vs After NFSU")
```

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

# Section 9: INGARCH Spatio-Temporal Model (`hhh4`)

## Background and Motivation

The GLM models in Sections 4–6 treat each week's shooting count as
depending on time-level covariates (trend, season, NFSU indicator) but
do not explicitly model the **contagion process** — the idea that a
shooting this week makes shootings more likely *in the same location*
next week. This is the self-excitation or "repeat victimization" concept
central to the Hawkes process literature.

The **INGARCH** (Integer-valued Generalized AutoRegressive Conditional
Heteroskedastic) model handles this directly. For each grid cell $i$ at
week $t$, we model:

$$Y_{i,t} \mid Y_{i,t-1} \sim \text{Poisson}(\lambda_{i,t})$$

The conditional mean $\lambda_{i,t}$ decomposes into two additive parts:

$$\lambda_{i,t} = \underbrace{\mu_i}_{\text{background rate}} + \underbrace{\phi \cdot Y_{i,t-1}}_{\text{repeat victimization}}$$

-   **Background rate** $\mu_i$: the baseline expected count for grid
    $i$, driven by structural features of the location (demographics,
    geography, persistent criminal activity). This is the "endemic"
    component.
-   **Repeat victimization** $\phi \cdot Y_{i,t-1}$: last week's count
    in the same grid multiplies forward into this week. $\phi$ is the
    self-excitation parameter — if $\phi > 0$, a spike in violence tends
    to sustain itself. This is the "autoregressive" component.

**The key research question:** Did the formation of the NFSU (January 1,
2024) reduce $\phi$? A significant drop in $\phi$ after the unit formed
would indicate the unit disrupted the contagion mechanism — places that
spiked were no longer staying spiked.

We fit this using `hhh4()` from the `surveillance` package, which is
purpose-built for this class of spatio-temporal count models.

------------------------------------------------------------------------

## 9.1 Load `surveillance` Package

```{r message=FALSE, warning=FALSE}
# The surveillance package provides hhh4(), which fits multivariate
# count time series models of the INGARCH / Hawkes process family.
# It requires data structured as an sts (surveillance time series) object.
library(surveillance)
```

------------------------------------------------------------------------

## 9.2 Build the `sts` Object

`hhh4()` does not accept a standard data frame. It requires a special
**`sts`** object where:

-   Rows = time periods (weeks)
-   Columns = spatial units (grid cells)
-   A neighbourhood adjacency matrix links grids to their neighbours

```{r}
# =============================================================
# STEP 1: Pivot panel from long to wide format
# =============================================================
# The panel data frame (from Section 2) has one row per week × grid.
# hhh4 needs it as a matrix: rows = weeks, columns = grids.
# Missing week × grid combos are already filled with 0 from the panel build.

# Grab the sorted unique weeks and grid IDs — we need these later
# for building the NFSU covariate matrix and aligning fitted values.
all_grid_ids  <- sort(unique(panel$grid_id))
all_week_vals <- sort(unique(panel$week))

# Pivot to wide: each column is one grid, each row is one week
count_wide <- panel %>%
  dplyr::select(week, grid_id, count) %>%
  pivot_wider(
    names_from  = grid_id,     # one column per grid
    values_from = count,
    values_fill = 0            # grids with no shootings that week = 0
  ) %>%
  arrange(week)                # make sure weeks are in chronological order

# Drop the week column and convert to a plain numeric matrix
count_matrix <- count_wide %>%
  dplyr::select(-week) %>%
  as.matrix()

# Label rows with their week dates — used later to align fitted/observed
rownames(count_matrix) <- as.character(count_wide$week)

cat("Count matrix dimensions:", nrow(count_matrix), "weeks x",
    ncol(count_matrix), "grids\n")
cat("Total shooting events in matrix:", sum(count_matrix), "\n")
cat("Overall zero proportion:", round(mean(count_matrix == 0), 3),
    "— very sparse; this is expected for weekly grid data\n")
```

```{r}
# =============================================================
# STEP 2: Build the neighbourhood adjacency matrix
# =============================================================
# hhh4 uses spatial adjacency to allow for spatial spread of events
# (the 'ne' or neighbourhood component). Even if we don't model spatial
# spread explicitly, hhh4 needs the adjacency matrix to be present.
#
# IMPORTANT: the adjacency matrix rows/columns must be in the SAME ORDER
# as the columns of count_matrix. We enforce this with arrange(match(...)).

grid_sf_ordered <- grid_sf %>%
  rename(grid_id = Name) %>%
  filter(grid_id %in% colnames(count_matrix)) %>%
  arrange(match(grid_id, colnames(count_matrix)))  # match column order exactly

# Queen contiguity: two grids are neighbours if they share any edge or corner.
# zero.policy = TRUE handles grids with no neighbours (island grids).
nb_hhh4 <- poly2nb(grid_sf_ordered, queen = TRUE)
adj_mat  <- nb2mat(nb_hhh4, style = "B", zero.policy = TRUE)  # binary 0/1 matrix
rownames(adj_mat) <- grid_sf_ordered$grid_id
colnames(adj_mat) <- grid_sf_ordered$grid_id

cat("Adjacency matrix:", nrow(adj_mat), "x", ncol(adj_mat), "\n")
cat("Average neighbours per grid:", round(mean(rowSums(adj_mat)), 2), "\n")
```

```{r}
# =============================================================
# STEP 3: Create the sts object
# =============================================================
# sts() packages the count matrix and neighbourhood matrix together.
# frequency = 52 tells hhh4 we have weekly data (52 weeks per year).

sts_obj <- sts(
  observed      = count_matrix,
  neighbourhood = adj_mat,
  frequency     = 52   # weekly periodicity
)

cat("sts object created successfully\n")
cat("Time points (weeks):", nrow(sts_obj), "\n")
cat("Spatial units (grids):", ncol(sts_obj), "\n")
```

------------------------------------------------------------------------

## 9.3 Build the NFSU Covariate Matrix

To allow $\phi$ to differ before vs after NFSU formation, we need to
pass an NFSU indicator as a covariate into the AR component of `hhh4`.

`hhh4()` requires time-varying covariates as **plain 2D matrices**
`[weeks × grids]` placed in the top-level `data` list of the control
object. The same value applies to all grids in a given week, so we
replicate the weekly indicator across all grid columns.

```{r}
# =============================================================
# NFSU indicator vector: 0 = before unit formed, 1 = after
# =============================================================
# The NFSU (Non-Fatal Shooting Unit) was formed January 1, 2024.
# We code weeks starting on or after that date as 1, prior weeks as 0.
nfsu_vec <- as.integer(all_week_vals >= as.Date("2024-01-01"))

cat("Weeks before NFSU (nfsu = 0):", sum(nfsu_vec == 0), "\n")
cat("Weeks after  NFSU (nfsu = 1):", sum(nfsu_vec == 1), "\n")

# =============================================================
# Replicate the weekly indicator across all grid columns
# =============================================================
# hhh4 expects a [weeks x grids] matrix for time-varying covariates.
# Since the NFSU effect is the same for all grids in a given week,
# we tile the vector horizontally to fill the matrix.
nfsu_matrix <- matrix(
  rep(nfsu_vec, ncol(count_matrix)),  # repeat the vector once per grid
  nrow = nrow(count_matrix),
  ncol = ncol(count_matrix)
)

cat("NFSU matrix dimensions:", nrow(nfsu_matrix), "weeks x",
    ncol(nfsu_matrix), "grids — ready for hhh4\n")
```

------------------------------------------------------------------------

## 9.4 Model 1 — Baseline: Global Background + Global Self-Excitation

**Equation:** $$\lambda_{i,t} = e^{\beta_0} + \phi \cdot Y_{i,t-1}$$

This is the simplest INGARCH model:

-   A single global intercept $e^{\beta_0}$ acts as the background rate
    (the same for all grids and all time periods — a strong
    simplification).
-   A single global $\phi$ captures how strongly last week's count in a
    grid predicts this week's count in the same grid.
-   No NFSU effect — this is the **null model** for the NFSU hypothesis.

In `hhh4()` terms:

-   `ar` = autoregressive component → captures $\phi \cdot Y_{i,t-1}$
-   `end` = endemic component → captures the background rate
    $e^{\beta_0}$

Note: `hhh4` works on the log scale internally, so `ar.1` and `end.1` in
the output are log-scale coefficients. To get $\phi$ on the natural
scale: $\phi = e^{\text{ar.1}}$.

```{r}
# =============================================================
# Define the control list for Model 1
# =============================================================
# f = ~ 1 means "fit an intercept only" for that component.
# family = "Poisson" specifies the count distribution.

ctrl_m1 <- list(
  ar     = list(f = ~ 1),   # global phi: one AR coefficient for all grids/times
  end    = list(f = ~ 1),   # global background: one intercept for all grids/times
  family = "Poisson"
)

# Fit Model 1 — wrap in tryCatch so errors don't halt the whole document
fit_m1 <- tryCatch(
  hhh4(sts_obj, control = ctrl_m1),
  error = function(e) {
    cat("Model 1 fitting error:", conditionMessage(e), "\n")
    NULL
  }
)

# Print the summary if the model converged
if (!is.null(fit_m1)) {
  cat("=== Model 1: Global Background + Global Self-Excitation ===\n")
  cat("Converged:", fit_m1$convergence, "\n\n")
  print(summary(fit_m1))
  cat("\nNote: ar.1 and end.1 are on the log scale.\n")
  cat("phi (self-excitation) = exp(ar.1) =",
      round(exp(coef(fit_m1)["ar.1"]), 4), "\n")
  cat("This means: a shooting this week raises next week's expected",
      "count in the same grid by", round(exp(coef(fit_m1)["ar.1"]) * 100, 1),
      "percentage points on average.\n")
}
```

------------------------------------------------------------------------

## 9.5 Model 2a — Key Model: Global Background + NFSU-Stratified Self-Excitation

**Equation:** $$\lambda_{i,t} = e^{\beta_0} + \phi_j \cdot Y_{i,t-1}$$

where $j = 0$ before NFSU, $j = 1$ after NFSU. So:

-   $\phi_{\text{before}} = e^{\beta_{\text{ar}}}$
-   $\phi_{\text{after}} = e^{\beta_{\text{ar}} + \beta_{\text{nfsu}}}$

This model adds exactly **one parameter** (the NFSU modifier on $\phi$)
to Model 1. If $\beta_{\text{nfsu}} < 0$, the self-excitation parameter
dropped after the NFSU formed — consistent with the unit disrupting the
contagion mechanism.

**Why this model is identifiable (unlike a version with grid random
effects):** We only have \~101 after-NFSU weeks spread across 200 sparse
grids. If we also fit 200 grid-specific intercepts simultaneously, there
isn't enough data per grid to identify both the grid effects AND the
NFSU AR modifier. By keeping the background rate global here, we use all
the data to estimate the one coefficient we care about
($\beta_{\text{nfsu}}$).

```{r}
# =============================================================
# Define the control list for Model 2a
# =============================================================
# ar formula: ~ 1 + nfsu
#   → baseline AR intercept (ar.1) + NFSU modifier (ar.nfsu)
#   → phi_before = exp(ar.1)
#   → phi_after  = exp(ar.1 + ar.nfsu)
#
# data = list(nfsu = nfsu_matrix):
#   → the NFSU covariate [205 weeks x 200 grids] lives at the TOP LEVEL
#     of the control list, NOT nested inside ar. This is an hhh4 quirk.
#
# start: warm-start from M1 estimates so the optimizer has a good
#   starting point, especially important for the global intercept.
#
# optimizer stop: increase max iterations from the default (10) to 300
#   to give the model room to converge on the additional parameter.

ctrl_m2a <- list(
  ar     = list(f = ~ 1 + nfsu),       # phi differs before vs after NFSU
  end    = list(f = ~ 1),              # single global background rate
  family = "Poisson",
  data   = list(nfsu = nfsu_matrix),   # NFSU indicator matrix — top-level!
  optimizer = list(
    stop = list(iter.max = 300, tol = 1e-6)  # more iterations than default
  ),
  start = list(
    fixed = c(
      "ar.1"    = coef(fit_m1)["ar.1"],   # start phi at M1 estimate
      "ar.nfsu" = 0,                       # start NFSU modifier at 0 (no effect)
      "end.1"   = coef(fit_m1)["end.1"]   # start background at M1 estimate
    )
  )
)

# suppressWarnings: hhh4 prints a convergence warning mid-run even when
# it converges — we check fit_m2a$convergence explicitly below instead.
fit_m2a <- suppressWarnings(tryCatch(
  hhh4(sts_obj, control = ctrl_m2a),
  error = function(e) {
    cat("Model 2a fitting error:", conditionMessage(e), "\n")
    NULL
  }
))

if (!is.null(fit_m2a)) {
  cat("=== Model 2a: Global Background + NFSU-Stratified AR ===\n")
  cat("Converged:", fit_m2a$convergence, "\n\n")
  print(summary(fit_m2a))

  # --- phi extraction (natural scale) ---
  cat("\n--- phi estimates (natural scale) ---\n")
  cat("Note: ar.1 and ar.nfsu are on the log scale internally.\n")
  cat("  phi_before = exp(ar.1)\n")
  cat("  phi_after  = exp(ar.1 + ar.nfsu)\n\n")

  coefs_m2a  <- fit_m2a$coefficients
  b0_m2a     <- coefs_m2a["ar.1"]
  b1_m2a     <- coefs_m2a["ar.nfsu"]
  phi_before_m2a <- exp(b0_m2a)
  phi_after_m2a  <- exp(b0_m2a + b1_m2a)

  cat(sprintf("phi_before (pre-NFSU) : %.4f\n", phi_before_m2a))
  cat(sprintf("phi_after  (post-NFSU): %.4f\n", phi_after_m2a))
  cat(sprintf("Change in phi (after - before): %.4f\n", phi_after_m2a - phi_before_m2a))
  cat(sprintf("Relative change: %.1f%%\n",
      (phi_after_m2a - phi_before_m2a) / phi_before_m2a * 100))
}
```

------------------------------------------------------------------------

## 9.6 Model 2b — Spatial Heterogeneity: Grid-Specific Background + Global AR

**Equation:** $$\lambda_{i,t} = \mu_i + \phi \cdot Y_{i,t-1}$$

This model asks a different question from M2a: do different grid cells
have structurally different baseline crime rates? The `ri(type = "iid")`
term fits one intercept per grid — 200 independent parameters — so each
grid's background rate is estimated from its own history.

This is NOT the primary model for the NFSU hypothesis (no NFSU term),
but it's important for understanding whether the spatial heterogeneity
we saw in the LISA maps (Section 3) is also present in the model fit.
The massive AIC improvement over M1 confirms it is.

```{r}
# =============================================================
# Define the control list for Model 2b
# =============================================================
# end formula: ~ -1 + ri(type = "iid")
#   → -1 removes the global intercept
#   → ri(type = "iid") adds one independent intercept per grid cell
#   → result: 200 grid-specific background rates (mu_i)
#
# ar formula: ~ 1 → single global phi (same for all grids/times)
#
# Note: We do NOT include NFSU here. The 200 grid intercepts + NFSU
# interaction is not identifiable with this data (too sparse). M2a
# handles the NFSU question; M2b handles the spatial question.

ctrl_m2b <- list(
  ar     = list(f = ~ 1),                    # global phi
  end    = list(f = ~ -1 + ri(type = "iid")), # one intercept per grid
  family = "Poisson",
  optimizer = list(
    stop = list(iter.max = 500, tol = 1e-5)  # grid-level model needs more iterations
  ),
  start = list(
    fixed = c("ar.1" = coef(fit_m1)["ar.1"]) # warm-start AR from M1
  )
)

fit_m2b <- suppressWarnings(tryCatch(
  hhh4(sts_obj, control = ctrl_m2b),
  error = function(e) {
    cat("Model 2b fitting error:", conditionMessage(e), "\n")
    NULL
  }
))

if (!is.null(fit_m2b)) {
  cat("=== Model 2b: Grid-Specific Background + Global AR ===\n")
  cat("Converged:", fit_m2b$convergence, "\n\n")

  # Don't print summary — it has 200+ coefficients (one per grid).
  # Instead, print just the AR coefficient and model fit statistics.
  cat("Global phi (ar.1, log scale):",
      round(fit_m2b$coefficients["ar.1"], 5), "\n")
  cat("phi (natural scale) = exp(ar.1) =",
      round(exp(fit_m2b$coefficients["ar.1"]), 4), "\n")
  cat("Log-likelihood:", round(fit_m2b$loglikelihood, 2), "\n")
  cat("AIC (manual):",
      round(-2 * fit_m2b$loglikelihood + 2 * length(fit_m2b$coefficients), 2), "\n")
  cat("\nNote: The massive improvement over M1 confirms strong spatial",
      "heterogeneity in background rates across grids.\n")
}
```

------------------------------------------------------------------------

## 9.7 Formal Model Comparison

We compare all three models on log-likelihood and AIC. Because M1 and
M2a are nested (M2a adds exactly one parameter to M1), we can also run a
**likelihood ratio test** (LRT) to formally test whether the NFSU effect
on $\phi$ is statistically significant.

```{r}
# =============================================================
# Model comparison table
# =============================================================
# Collect all fitted models that ran successfully into a named list.
# This makes the comparison code robust — if a model failed, it's
# simply excluded from the table.

models_list <- list(
  "M1: Global background + global φ"    = fit_m1,
  "M2a: Global background + NFSU φ"     = fit_m2a,
  "M2b: Grid backgrounds + global φ"    = fit_m2b
)
models_list <- Filter(Negate(is.null), models_list)  # drop any NULLs

# Build comparison data frame.
# AIC = -2 * logLik + 2 * k, where k = number of parameters.
# We compute manually because AIC() may return NA if convergence is FALSE.
model_comp <- data.frame(
  Model     = names(models_list),
  Converged = sapply(models_list, function(m) m$convergence),
  LogLik    = sapply(models_list, function(m) round(m$loglikelihood, 2)),
  Params    = sapply(models_list, function(m) length(m$coefficients)),
  AIC       = sapply(models_list, function(m)
    round(-2 * m$loglikelihood + 2 * length(m$coefficients), 2)),
  row.names = NULL
)

knitr::kable(
  model_comp,
  caption = paste(
    "INGARCH Model Comparison.",
    "M2a vs M1: tests whether NFSU changed self-excitation (1 extra parameter).",
    "M2b vs M1: tests whether grid-specific backgrounds improve fit (198 extra parameters)."
  )
)
```

```{r}
# =============================================================
# Likelihood Ratio Test: M1 vs M2a
# =============================================================
# The LRT compares two nested models. M2a adds exactly 1 parameter
# (ar.nfsu) to M1. Under H0 (ar.nfsu = 0, no NFSU effect), the test
# statistic 2*(logLik(M2a) - logLik(M1)) follows a chi-squared
# distribution with 1 degree of freedom.
#
# A significant result means the NFSU modifier on phi is statistically
# supported — the self-excitation parameter did change after the unit formed.

if (!is.null(fit_m1) && !is.null(fit_m2a)) {

  # Compute the LRT statistic manually
  lrt_stat <- 2 * (fit_m2a$loglikelihood - fit_m1$loglikelihood)
  lrt_df   <- length(fit_m2a$coefficients) - length(fit_m1$coefficients)
  lrt_p    <- pchisq(lrt_stat, df = lrt_df, lower.tail = FALSE)

  cat("=== Likelihood Ratio Test: M1 vs M2a ===\n")
  cat("H0: NFSU has no effect on phi (ar.nfsu = 0)\n")
  cat("H1: NFSU changes phi (ar.nfsu ≠ 0)\n\n")
  cat("LRT statistic (chi-squared):", round(lrt_stat, 4), "\n")
  cat("Degrees of freedom:", lrt_df, "\n")
  cat("p-value:", format(lrt_p, digits = 4), "\n\n")

  if (lrt_p < 0.001) {
    cat("RESULT: Highly significant (p < 0.001).",
        "Strong evidence that phi changed after NFSU formation.\n")
  } else if (lrt_p < 0.05) {
    cat("RESULT: Significant (p < 0.05).",
        "Evidence that phi changed after NFSU formation.\n")
  } else {
    cat("RESULT: Not significant (p >= 0.05).",
        "Insufficient evidence of a change in phi after NFSU.\n")
  }
}
```

------------------------------------------------------------------------

## 9.8 Key Results — NFSU Effect on Self-Excitation ($\phi$)

This section translates the M2a log-scale coefficients into
interpretable $\phi$ values on the natural scale, with 95% confidence
intervals computed via the delta method (standard in `hhh4`).

```{r}
# =============================================================
# Extract phi_before and phi_after with confidence intervals
# =============================================================
# hhh4 stores coefficients on the log scale. To get phi on the
# natural scale we exponentiate:
#   phi_before = exp(ar.1)
#   phi_after  = exp(ar.1 + ar.nfsu)
#
# For confidence intervals, we use the coefficient standard errors
# from the model summary. Because phi = exp(linear combination),
# we build CIs on the log scale first, then exponentiate:
#   CI for phi_before: exp(ar.1 ± 1.96 * SE(ar.1))
#   CI for phi_after:  exp((ar.1 + ar.nfsu) ± 1.96 * sqrt(Var(ar.1 + ar.nfsu)))
#
# Var(ar.1 + ar.nfsu) = Var(ar.1) + Var(ar.nfsu) + 2*Cov(ar.1, ar.nfsu)

if (!is.null(fit_m2a) && fit_m2a$convergence) {

  # Pull the full coefficient vector
  coefs <- fit_m2a$coefficients
  b0    <- coefs["ar.1"]      # log-scale AR intercept
  b1    <- coefs["ar.nfsu"]   # log-scale NFSU modifier

  # Pull the variance-covariance matrix for the fixed effects
  # fitm2a$cov is the full VCV; we extract rows/cols for the AR terms
  vcv       <- fit_m2a$cov
  var_b0    <- vcv["ar.1",    "ar.1"]
  var_b1    <- vcv["ar.nfsu", "ar.nfsu"]
  cov_b0b1  <- vcv["ar.1",    "ar.nfsu"]

  # Standard errors on log scale
  se_before <- sqrt(var_b0)
  se_after  <- sqrt(var_b0 + var_b1 + 2 * cov_b0b1)

  # Point estimates on natural scale
  phi_before <- exp(b0)
  phi_after  <- exp(b0 + b1)

  # 95% confidence intervals (exponentiated from log-scale CIs)
  z95 <- qnorm(0.975)   # 1.96
  phi_before_lo <- exp(b0         - z95 * se_before)
  phi_before_hi <- exp(b0         + z95 * se_before)
  phi_after_lo  <- exp(b0 + b1    - z95 * se_after)
  phi_after_hi  <- exp(b0 + b1    + z95 * se_after)

  # Build a clean summary table
  phi_table <- data.frame(
    Period       = c("Before NFSU (2022–2023)", "After NFSU (2024–2025)"),
    phi_estimate = round(c(phi_before, phi_after), 4),
    CI_lower     = round(c(phi_before_lo, phi_after_lo), 4),
    CI_upper     = round(c(phi_before_hi, phi_after_hi), 4),
    Interpretation = c(
      paste0("A shooting raises next week's expected count by ~",
             round(phi_before * 100, 1), "pp in the same grid"),
      paste0("A shooting raises next week's expected count by ~",
             round(phi_after * 100, 1), "pp in the same grid")
    )
  )

  knitr::kable(
    phi_table,
    col.names = c("Period", "φ estimate", "95% CI lower",
                  "95% CI upper", "Plain-language interpretation"),
    caption   = paste(
      "Self-excitation parameter φ before and after NFSU formation.",
      "φ represents the fraction of this week's shootings in a grid",
      "attributable to last week's shootings in the same grid.",
      "CIs computed via delta method on the log scale."
    )
  )

  # Summary statistics for easy reference
  cat("\n=== Summary ===\n")
  cat("phi before NFSU:", round(phi_before, 4),
      sprintf("(95%% CI: %.4f – %.4f)\n", phi_before_lo, phi_before_hi))
  cat("phi after  NFSU:", round(phi_after, 4),
      sprintf("(95%% CI: %.4f – %.4f)\n", phi_after_lo, phi_after_hi))
  cat("Absolute reduction in phi:", round(phi_before - phi_after, 4), "\n")
  cat("Relative reduction:", round((1 - phi_after / phi_before) * 100, 1),
      "% drop in self-excitation after NFSU\n")

  # Final interpretation
  cat("\n=== Interpretation ===\n")
  if (phi_after < phi_before && phi_after_hi < phi_before_lo) {
    cat("The confidence intervals for phi_before and phi_after do NOT overlap.\n")
    cat("This provides strong evidence that self-excitation decreased\n")
    cat("significantly after the NFSU was formed — the LRT confirms this.\n")
    cat("A shooting in a grid cell was much less likely to 'stay hot'\n")
    cat("in the following week after the unit began operating.\n")
  } else if (phi_after < phi_before) {
    cat("phi decreased after NFSU, and the LRT is significant, but\n")
    cat("the confidence intervals overlap slightly — interpret with caution.\n")
  } else {
    cat("phi did not meaningfully decrease after NFSU formation.\n")
  }

} else if (!is.null(fit_m2a)) {
  # Model ran but didn't converge — we can still report raw coefficients
  # but note that the standard errors may be unreliable.
  cat("WARNING: M2a did not fully converge. SEs may be unreliable.\n")
  b0 <- fit_m2a$coefficients["ar.1"]
  b1 <- fit_m2a$coefficients["ar.nfsu"]
  cat("phi before (approx):", round(exp(b0), 4), "\n")
  cat("phi after  (approx):", round(exp(b0 + b1), 4), "\n")
} else {
  cat("M2a did not run. Cannot report phi estimates.\n")
}
```

------------------------------------------------------------------------

## 9.9 Fitted vs Observed — Model 1

A visual check of how well each model tracks the observed weekly counts
across all grids combined.

Note: `fitted()` returns one fewer row than `observed()` because the
AR(1) term requires $Y_{i,t-1}$, so the first week has no fitted value.
We align by trimming `observed` from the front.

```{r fig.height=5}
if (!is.null(fit_m1)) {

  # Get fitted values (sum across all grids to get city-wide weekly total)
  fit_weekly <- rowSums(fitted(fit_m1))
  n_fit      <- length(fit_weekly)

  # Trim observed and week labels to match the length of fitted
  # (AR lag means first week has no prediction)
  obs_weekly <- rowSums(observed(sts_obj))[
    (nrow(count_matrix) - n_fit + 1):nrow(count_matrix)
  ]
  weeks_seq <- as.Date(rownames(count_matrix))[
    (nrow(count_matrix) - n_fit + 1):nrow(count_matrix)
  ]

  diag_df <- data.frame(week = weeks_seq, observed = obs_weekly, fitted = fit_weekly)

  ggplot(diag_df, aes(x = week)) +
    geom_line(aes(y = observed, color = "Observed"), linewidth = 0.8) +
    geom_line(aes(y = fitted, color = "Fitted M1"), linewidth = 0.8,
              linetype = "dashed") +
    geom_vline(xintercept = as.Date("2024-01-01"),
               color = "red", linetype = "dotted") +
    annotate("text", x = as.Date("2024-01-01"),
             y = max(obs_weekly, na.rm = TRUE),
             label = "NFSU start", color = "red", hjust = -0.1, size = 3) +
    scale_color_manual(
      values = c("Observed" = "steelblue", "Fitted M1" = "darkorange")
    ) +
    labs(
      title    = "Model 1: Weekly Observed vs Fitted (Global Background + Global φ)",
      subtitle = "M1 uses a single global background and self-excitation rate across all grids",
      x = "Week", y = "Total Shootings (all grids)", color = ""
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

------------------------------------------------------------------------

## 9.10 Fitted vs Observed — Model 2a

```{r fig.height=5}
if (!is.null(fit_m2a)) {

  fit2_weekly <- rowSums(fitted(fit_m2a))
  n_fit2      <- length(fit2_weekly)

  obs_weekly <- rowSums(observed(sts_obj))[
    (nrow(count_matrix) - n_fit2 + 1):nrow(count_matrix)
  ]
  weeks_seq <- as.Date(rownames(count_matrix))[
    (nrow(count_matrix) - n_fit2 + 1):nrow(count_matrix)
  ]

  diag_df2 <- data.frame(week = weeks_seq, observed = obs_weekly, fitted = fit2_weekly)

  ggplot(diag_df2, aes(x = week)) +
    geom_line(aes(y = observed, color = "Observed"), linewidth = 0.8) +
    geom_line(aes(y = fitted, color = "Fitted M2a"), linewidth = 0.8,
              linetype = "dashed") +
    geom_vline(xintercept = as.Date("2024-01-01"),
               color = "red", linetype = "dotted") +
    annotate("text", x = as.Date("2024-01-01"),
             y = max(obs_weekly, na.rm = TRUE),
             label = "NFSU start", color = "red", hjust = -0.1, size = 3) +
    scale_color_manual(
      values = c("Observed" = "steelblue", "Fitted M2a" = "darkgreen")
    ) +
    labs(
      title    = "Model 2a: Weekly Observed vs Fitted (Global Background + NFSU φ)",
      subtitle = "M2a allows φ to differ before vs after NFSU formation",
      x = "Week", y = "Total Shootings (all grids)", color = ""
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

------------------------------------------------------------------------

## 9.10b Fitted vs Observed — Model 2b

```{r fig.height=5}
if (!is.null(fit_m2b)) {

  fit2b_weekly <- rowSums(fitted(fit_m2b))
  n_fit2b      <- length(fit2b_weekly)

  obs_weekly_2b <- rowSums(observed(sts_obj))[
    (nrow(count_matrix) - n_fit2b + 1):nrow(count_matrix)
  ]
  weeks_seq_2b <- as.Date(rownames(count_matrix))[
    (nrow(count_matrix) - n_fit2b + 1):nrow(count_matrix)
  ]

  diag_df2b <- data.frame(week = weeks_seq_2b,
                           observed = obs_weekly_2b,
                           fitted   = fit2b_weekly)

  ggplot(diag_df2b, aes(x = week)) +
    geom_line(aes(y = observed, color = "Observed"), linewidth = 0.8) +
    geom_line(aes(y = fitted, color = "Fitted M2b"), linewidth = 0.8,
              linetype = "dashed") +
    geom_vline(xintercept = as.Date("2024-01-01"),
               color = "red", linetype = "dotted") +
    annotate("text", x = as.Date("2024-01-01"),
             y = max(obs_weekly_2b, na.rm = TRUE),
             label = "NFSU start", color = "red", hjust = -0.1, size = 3) +
    scale_color_manual(
      values = c("Observed" = "steelblue", "Fitted M2b" = "purple")
    ) +
    labs(
      title    = "Model 2b: Weekly Observed vs Fitted (Grid-Specific Background + Global φ)",
      subtitle = "M2b uses grid-specific background rates (ri) and a single global phi",
      x = "Week", y = "Total Shootings (all grids)", color = ""
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

------------------------------------------------------------------------

## 9.11 Residual Diagnostics — ACF/PACF

If the model has successfully captured the temporal autocorrelation in
the data, the ACF and PACF of the Pearson residuals should show no
significant bars beyond the dashed confidence bounds.

Pearson residuals:
$(Y_{i,t} - \hat{\lambda}_{i,t}) / \sqrt{\hat{\lambda}_{i,t}}$

We average these across all grids within each week to get a single
residual time series for visual inspection. Plots are shown for all
three spatio-temporal models so their residual autocorrelation can be
compared directly.

### ACF/PACF — Model 1

```{r fig.height=4}
if (!is.null(fit_m1)) {

  fit_mat_m1 <- fitted(fit_m1)
  n_fit_m1   <- nrow(fit_mat_m1)

  obs_mat_m1 <- observed(sts_obj)[
    (nrow(count_matrix) - n_fit_m1 + 1):nrow(count_matrix), ]

  pearson_mat_m1  <- (obs_mat_m1 - fit_mat_m1) / sqrt(fit_mat_m1 + 1e-8)
  resid_weekly_m1 <- rowMeans(pearson_mat_m1, na.rm = TRUE)

  op <- par(no.readonly = TRUE)
  on.exit(par(op), add = TRUE)
  par(mfrow = c(1, 2))
  acf(resid_weekly_m1,
      main    = "ACF — M1 Weekly Pearson Residuals",
      lag.max = 30,
      xlab    = "Lag (weeks)")
  pacf(resid_weekly_m1,
       main    = "PACF — M1 Weekly Pearson Residuals",
       lag.max = 30,
       xlab    = "Lag (weeks)")
  par(op)

  cat("=== M1 ACF/PACF Interpretation ===\n")
  cat("M1 uses a single global background and a single global phi.\n")
  cat("Bars within dashed blue lines: no residual autocorrelation at that lag.\n")
  cat("Persistent bars suggest M1 is not fully capturing temporal dynamics.\n")
}
```

### ACF/PACF — Model 2a

```{r fig.height=4}
if (!is.null(fit_m2a)) {

  fit_mat_m2a <- fitted(fit_m2a)
  n_fit_m2a   <- nrow(fit_mat_m2a)

  obs_mat_m2a <- observed(sts_obj)[
    (nrow(count_matrix) - n_fit_m2a + 1):nrow(count_matrix), ]

  pearson_mat_m2a  <- (obs_mat_m2a - fit_mat_m2a) / sqrt(fit_mat_m2a + 1e-8)
  resid_weekly_m2a <- rowMeans(pearson_mat_m2a, na.rm = TRUE)

  op <- par(no.readonly = TRUE)
  on.exit(par(op), add = TRUE)
  par(mfrow = c(1, 2))
  acf(resid_weekly_m2a,
      main    = "ACF — M2a Weekly Pearson Residuals",
      lag.max = 30,
      xlab    = "Lag (weeks)")
  pacf(resid_weekly_m2a,
       main    = "PACF — M2a Weekly Pearson Residuals",
       lag.max = 30,
       xlab    = "Lag (weeks)")
  par(op)

  cat("=== M2a ACF/PACF Interpretation ===\n")
  cat("M2a allows phi to differ before vs after NFSU formation.\n")
  cat("Bars within dashed blue lines: no residual autocorrelation at that lag.\n")
  cat("Bars exceeding the dashed lines: remaining temporal structure the model\n")
  cat("hasn't captured — could suggest a higher-order AR term or missed covariate.\n")
}
```

### ACF/PACF — Model 2b

```{r fig.height=4}
if (!is.null(fit_m2b)) {

  fit_mat_m2b <- fitted(fit_m2b)
  n_fit_m2b   <- nrow(fit_mat_m2b)

  obs_mat_m2b <- observed(sts_obj)[
    (nrow(count_matrix) - n_fit_m2b + 1):nrow(count_matrix), ]

  pearson_mat_m2b  <- (obs_mat_m2b - fit_mat_m2b) / sqrt(fit_mat_m2b + 1e-8)
  resid_weekly_m2b <- rowMeans(pearson_mat_m2b, na.rm = TRUE)

  op <- par(no.readonly = TRUE)
  on.exit(par(op), add = TRUE)
  par(mfrow = c(1, 2))
  acf(resid_weekly_m2b,
      main    = "ACF — M2b Weekly Pearson Residuals",
      lag.max = 30,
      xlab    = "Lag (weeks)")
  pacf(resid_weekly_m2b,
       main    = "PACF — M2b Weekly Pearson Residuals",
       lag.max = 30,
       xlab    = "Lag (weeks)")
  par(op)

  cat("=== M2b ACF/PACF Interpretation ===\n")
  cat("M2b uses grid-specific background rates (ri) with a global phi.\n")
  cat("If M2b residuals show less autocorrelation than M1/M2a, the spatial\n")
  cat("random effects are absorbing temporal structure as well.\n")
  cat("Bars within dashed blue lines: no residual autocorrelation at that lag.\n")
}
```

------------------------------------------------------------------------

## 9.12 Important Caveats

Before presenting these results, note the following limitations:

1.  **Quasi-experimental design**: The before/after contrast cannot rule
    out other changes that occurred simultaneously with NFSU formation
    in January 2024 (e.g., seasonal effects in January, other policing
    changes, demographic shifts). The model controls for seasonality via
    the GLM stage but not within the INGARCH model itself.

2.  **Zero inflation**: 93.8% of the weekly grid × time cells are zero.
    The Poisson INGARCH model handles sparse counts but may underfit
    grids that were consistently zero throughout. Restricting to the
    top-quartile grids (as in Section 6) addresses this.

3.  **Spatial component**: These models do not include the neighbourhood
    (`ne`) component of `hhh4`, which would model spatial spread of
    violence across neighbouring grids. The Moran's I on residuals
    (Section 6) showed significant residual spatial structure,
    suggesting a full spatio-temporal model (with `ne`) could improve
    fit further.

4.  **Identifiability of M2b+NFSU**: Combining 200 grid random effects
    with an NFSU interaction in AR is not identifiable with this dataset
    (as we found during fitting). M2a cleanly isolates the NFSU question
    and is the appropriate model for inference.
